---
title: GATA3 ChIP-seq analysis 
author:
- Siyu Sun
- Michael J. Guertin
header-includes:
- \usepackage{color}
- \usepackage{float}
- \DeclareUnicodeCharacter{2212}{-}
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    toc: true
fontsize: 14pt
geometry: margin=1in
---


# ChIP-seq experiments

ChIP-seq measures protein abundance on DNA and one can quantify
protein modifications if an antibody to the modification is
available. 

# ChIP-seq experimental design

Controls are important.  An IgG (or mock) IP is superior to input as
a control. I am happy to discuss the design of ChIP-seq experiments 
with you **before** you generate any data. It is much easier to design
the proper experiments and controls and analyze the data compared to
the alternative of trying to salvage poorly controlled experiments.

# Info about this experiment

MCF7 cells with the dTAG-GATA3 (clone522) grow in charcoal stripped media (ER deprived) are treated under 3 conditions: 1) CC: null+null; 2) CE: null+ E2; 3) dE: dTAG+E2. \
Cells are ChIPed with IgG, CTCF, GATA3, ER, respectively; plus an parallel-ChiP of CTCF and ER.

# ChIP-seq analysis

First we name the experiments with the trailing name being `PE1` or
`PE` for paired end data. 

```{r engine='bash', eval=F, echo=TRUE}
for i in *_S*R1_001.fastq.gz
do
    nm=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_S" '{print $1}')
    echo $nm
	two=$(echo $i | awk -F"/" '{print $NF}' | awk -F"R1_001.fastq.gz" '{print $1}')
	mv $i ${nm}_PE1.fastq.gz
	mv ${two}R2_001.fastq.gz ${nm}_PE2.fastq.gz
done
```


## Cut off the adapter with `cutadapt`

In our daily workflow in the lab we use
`cutadapt` to remove adapeter sequences. The options we use below are `-j` for the number of cores
to use, `-m` specifies the minimal lenght of a read to keep after
adapter sequence removal, and `-O` is the number of bases to trim off the
end of the read if it overlaps with the adapter sequence. If the
genome is 25% of each base, then you would expect one quarter of the
reads that have no adapter to have the trailing base
trimmed. Likewise, approximately 1/16 of the remaining
reads without the adapter will have the final two bases
trimmed. Technically these values are not exact, because the reads with
matches to longer trailing k-mers (in this case 19-mers) would be
removed first, then 18-mer matches removed, etcetera... The `-a` and `-A`
options are the adapter sequences of the PE1 and PE2 reads.  The
output file is `-o` (PE1) and `-p` PE2. The last two positional
areguments are the input `fastq` files. Of course we save the output
to a log file


```{r engine='bash', eval=F, echo=TRUE}
module load cutadapt
for i in *PE1.fastq.gz
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1" '{print $1}')
	echo $name
	echo unzipping $i
	gunzip $i
	echo unzipping ${name}_PE2.fastq.gz
	gunzip ${name}_PE2.fastq.gz
	cutadapt -a AGATCGGAAGAGCACACGTCTGAACTCCAGTCA -A AGATCGGAAGAGCGTCGTGTAGGGAAAGAGTGT -j 8 -m 10 -O 1 -o ${name}_PE1_no_adapt.fastq -p ${name}_PE2_no_adapt.fastq ${name}_PE1.fastq ${name}_PE2.fastq 2>&1 | tee ${name}_cutadapt.log
done
```

## Align to the human genome

Now we are aligning to the `hg38.fa` genome just as we have previously
for ChIP-seq, the difference is that the libraries are paired end. Note the `-1` and
`-2` options for the respective paired-end `fastq` files. There is no
need to save the output `sam` file, so the output is piped to
`samtools` to convert to `bam`, then sorted by name (`-n`) so paired
end reads are adjacent in the file, then piped to `samtools fixmate`
which adds information about the fragment length by comparing the PE1
and PE2 coordinates, then the files are sorted by coordinate, then
piped to `samtools markdup` to remove duplicate reads. Duplicate reads
have the same PE1 and PE2 ends. This is very unlikely to happen by
chance unless you sequence to very high read depth, so these reads are
considered PCR amplicon duplicates. The `fixmate` step is necessary to
pipe to `markdup`. We use the entire DNA fragment for making the
browser tracks, which is accomplished by converted the paired-end
`bam` file to a `bed12` and taking out discordant reads. We also
normalize to read depth. \


First we are getting reference genome and build the genome index \
```{r engine='bash', eval=F, echo=TRUE}
#! /bin/sh

#SBATCH --job-name=getrefgenome.sh     # name for job
#SBATCH -N 1
#SBATCH -n 1
#SBATCH -c 32
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=32G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o getrefgenome.sh_%j.out
#SBATCH -e getrefgenome.sh_%j.err

module load genometools/1.5.10
module load bowtie2
export PATH=$PATH:/home/FCAM/ssun/packages

wget https://hgdownload.cse.ucsc.edu/goldenpath/hg38/bigZips/hg38.fa.gz
gunzip hg38.fa.gz
bowtie2-build hg38.fa hg38

wget https://hgdownload-test.gi.ucsc.edu/goldenPath/hg38/bigZips/hg38.chrom.sizes
```

This is the chip_alignment_230718.sh file content: \
```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=chip_alignment_230718.sh     # name for job
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 32                  
#SBATCH -p general           
#SBATCH --qos=general       
#SBATCH --mem=32G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o chip_alignment_230718.sh_%j.out
#SBATCH -e chip_alignment_230718.sh_%j.err

hostname

name=230718

# I have seqOutBias and other useful tools export to my Path
export PATH=$PATH:/home/FCAM/ssun/packages/

module load samtools/1.16.1
module load genometools/1.5.10
module load ucsc_genome/2012.05.22
module load rust
module load bowtie2
module load bedtools


sizes=/home/FCAM/ssun/Genome/hg38.chrom.sizes

genome=/home/FCAM/ssun/Genome/hg38.fa
genome_index=/home/FCAM/ssun/Genome/hg38_bt2/hg38
ncore=30 
tallymer=hg38.tal_42.gtTxt.gz

bowtie2 -p $ncore --maxins 800 -x $genome_index -1 ${name}_PE1_no_adapt.fastq -2 ${name}_PE2_no_adapt.fastq | samtools sort -@ $ncore -n -o ${name}.bw.bam
gzip ${name}_PE1_no_adapt.fastq
gzip ${name}_PE2_no_adapt.fastq
gzip ${name}_PE2.fastq
gzip ${name}_PE1.fastq
samtools fixmate -m ${name}.bw.bam - | samtools sort -@ $ncore - | samtools markdup -s -r - ${name}.hg38.bam
seqOutBias ${genome} ${name}.hg38.bam --shift-counts --no-scale \
                                      --bw=${name}.bigWig --read-size=42 --tallymer=$tallymer 2>&1 | tee ${name}_seqOutBias.log
samtools sort -@ $ncore -n -o ${name}.sorted.bam ${name}.hg38.bam
# the above samtools -n is sorting files by queryname, which is important for the next bedtools bamtobed command with the -bedpe flag; 
# the bedtools bamtobed -bedpe is generating warning signs:
# *****WARNING: Query {...} is marked as paired, but its mate does not occur next to it in your BAM file.  Skipping.
# This might due to some single end sequences, I checked the fraction of these not paired sequences, it is <1.5%, so we don't need to worry.
bedtools bamtobed -i ${name}.sorted.bam -bedpe > ${name}_bed12.bed

awk '$1==$4 {print $0}' ${name}_bed12.bed | awk '{OFS="\t";} {print $1, $2, $6}' | awk '$1!="." && $3>$2 && (($3 - $2)<2000) {print $0}' | sort -k1,1 -k2,2n > ${name}_read_span.bed
genomeCoverageBed -bg -i ${name}_read_span.bed -g $sizes > ${name}.bedGraph
depth=`awk -F'\t' '{sum+=$5;}END{print sum;}' ${name}.hg38_not_scaled.bed`
scaled=$(bc <<< "scale=3 ; 10000000 / $depth")
echo $scaled
awk -v scaled="$scaled" '{OFS="\t";} {print $1, $2, $3, $4*scaled}' ${name}.bedGraph > ${name}_normalized.bedGraph
wigToBigWig -clip ${name}_normalized.bedGraph $sizes ${name}_normalized.bigWig
```

## Run the previous chunk in parallel 

generating tallymer files use seqOutBias:
```{r engine='bash', eval=F, echo=TRUE}
#Compute mappability for the given read length and the k-mer that corresponds to each possible read alignment position
#time-consuming but only need to run this one time
seqOutBias seqtable hg38.fa --read-size=42
```


```{r engine='bash', eval=F, echo=TRUE}
file=chip_alignment_230718.sh 

for i in *_PE1.fastq
do
    nm=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1.fastq" '{print $1}')
    fq=$(echo $i | rev | cut -f 1 -d '/' | rev)
    echo $nm
    echo $fq
    sed -e "s/230718/${nm}/g" "$file" > chip_alignment_${nm}.sh
    sbatch chip_alignment_${nm}.sh
    sleep 1
done

```


# Combining replicates for the genome browser

Usually we combine replicates into a single track for visualization
and we compare the tracks between conditions. It is important that we
read-depth normalize before we combine the signal from each
replicate. Otherwise, we would be weighting replicates differently;
for example, if a library is sequenced to twice the read depth and we
combine first then read depth normalize, then the more high coverage
data is weighted twice as much in the final visualization. All the
relevant track files are here:
[http://guertinlab.cam.uchc.edu/znf143_hub/](http://guertinlab.cam.uchc.edu/znf143_hub/)

```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=combine_rep.sh     # name for job
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 32                  
#SBATCH -p general           
#SBATCH --qos=general       
#SBATCH --mem=32G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o combine_rep.sh_%j.out
#SBATCH -e combine_rep.sh_%j.err

module load ucsc_genome/2012.05.22
sizes=/home/FCAM/ssun/Genome/hg38.chrom.sizes

for i in *MCF*rep1_PE1.fastq.gz
do
    nm=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1.fastq.gz" '{print $1}')
    name=$(echo $nm | awk -F"_rep1" '{print $1}')
    echo $name
    reps=$(ls ${name}_rep*normalized.bigWig | wc -w | bc)
	echo $reps
    files=$(ls ${name}_rep*normalized.bigWig)
	echo $files
    bigWigMerge $files tmp.bg
    scaleall=$(bc <<< "scale=4 ; 1.0 / $reps")
    echo scale:
    echo $scaleall
    awk -v scaleall="$scaleall" '{OFS="\t";} {print $1, $2, $3, $4*scaleall}' tmp.bg > ${name}_normalized.bedGraph
	rm tmp.bg
	wigToBigWig ${name}_normalized.bedGraph $sizes ${name}.bigWig
    awk -v var="$name" 'BEGIN {  print "browser position chr11:5,289,521-5,291,937"; print "track type=bedGraph name=\"" var "\" description=\"" var "_bedGraph\" visibility=full autoScale=on alwaysZero=on color=0,0,0"}  { print $0}' ${name}_normalized.bedGraph > ${name}_header_normalized.bedGraph
    gzip ${name}_header_normalized.bedGraph
done

# side note: only few files has replicates: CTCF_CC (2 reps), ER_CE (2 reps), and GATA_CC (3reps)
```

we have single replicate for IgG_dE, IgG_CE, and IgG_CC. Here I am trying to merge the three IgG files together
```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
module load ucsc_genome/2012.05.22
sizes=/home/FCAM/ssun/Genome/hg38.chrom.sizes
name=MCF7_dTAGGATA522_IgG
reps=3

files=$(ls ${name}*_normalized.bigWig)
echo $files

bigWigMerge $files tmp.bg
scaleall=$(bc <<< "scale=4 ; 1.0 / $reps")
echo scale:
echo $scaleall
awk -v scaleall="$scaleall" '{OFS="\t";} {print $1, $2, $3, $4*scaleall}' tmp.bg > ${name}_normalized.bedGraph
rm tmp.bg
wigToBigWig ${name}_normalized.bedGraph $sizes ${name}.bigWig
awk -v var="$name" 'BEGIN {  print "browser position chr11:5,289,521-5,291,937"; print "track type=bedGraph name=\"" var "\" description=\"" var "_bedGraph\" visibility=full autoScale=on alwaysZero=on color=0,0,0"}  { print $0}' ${name}_normalized.bedGraph > ${name}_header_normalized.bedGraph
gzip ${name}_header_normalized.bedGraph

```


```{r engine='bash', eval=F, echo=TRUE}
for i in *MCF*PE1.fastq.gz
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1.fastq.gz" '{print $1}')
    mkdir ${name}_files
	  mv ${name}* ./${name}_files
done
```


# Peak calling

```{r engine='bash', eval=F, echo=TRUE}
#! /bin/sh
#SBATCH --job-name=chip_peak_calling.sh     # name for job
#SBATCH -N 1
#SBATCH -n 1
#SBATCH -c 2
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=32G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=guertin@uchc.edu
#SBATCH -o chip_peak_calling.sh_%j.out
#SBATCH -e chip_peak_calling.sh_%j.err

hostname

module load macs3

mkdir temp_macs
macs3 callpeak --call-summits -t *_CTCF_*bam -c *IgG*bam -n CTCF_ChIP -g hs -q 0.01 --keep-dup all -f BAMPE --nomodel --tempdir temp_macs
macs3 callpeak --call-summits -t *cont_HA_*bam -c *IgG*bam -n ZNF143_ChIP -g hs -q 0.01 --keep-dup all -f BAMPE --nomodel --tempdir temp_macs
```

## Removing peaks on contigs and within blacklisted regions


You can Google "blacklisted genomic regions" to find a
set of region in the genome in `bed` format that have an over-representation of
reads regardless of the experiment. Recall this is even more necessary
because we do not have a control data set. We can also remove peaks on non-canonical chromosomes with
`grep -v`. 

```{r engine='bash', eval=F, echo=TRUE}
module load deeptools/3.5.0
module load bedtools
wget https://github.com/Boyle-Lab/Blacklist/raw/master/lists/hg38-blacklist.v2.bed.gz
gunzip hg38-blacklist.v2.bed.gz
blacklist=hg38-blacklist.v2.bed
sizes=/home/FCAM/meds5420/genomes/hg38.chrom.sizes


for i in *_summits.bed
do
	name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_summits.bed" '{print $1}')
	echo $name
	grep -v "random" ${name}_summits.bed | grep -v "chrUn" | grep -v "chrEBV" | grep -v "chrM" | grep -v "alt" | intersectBed -v -a - -b $blacklist > ${name}_summits_final.bed
	slopBed -b 70 -i ${name}_summits_final.bed -g $sizes  | sort -k1,1 -k2,2n > ${name}_summit_140window.bed
   	slopBed -b 200 -i ${name}_summits_final.bed -g $sizes  | sort -k1,1 -k2,2n > ${name}_summit_400window.bed
done

```

# Size factors from library 

```{r engine='bash', eval=F, echo=TRUE}

#size factors

module load samtools/1.12
for i in HA
do
  echo $i
  > ${i}_header.txt
  > ${i}_reads.txt
  for j in HEK_ZNF-dTAG*_${i}_*.bam
  do
    echo $j
    name=$(echo $j | awk -F".sorted.bam" '{print $1}')
    echo $name | paste ${i}_header.txt - > ${i}_tmp.txt 
    mv ${i}_tmp.txt ${i}_header.txt
    reads=`samtools view -c -f 0x3 $j`
    echo $reads | paste ${i}_reads.txt - > ${i}_tmp.txt 
    mv ${i}_tmp.txt ${i}_reads.txt
  done  
  cat ${i}_header.txt ${i}_reads.txt > ${i}_tmp.txt
  mv ${i}_tmp.txt ${i}_reads.txt
  rm ${i}_header.txt
done 

```

## Differential binding and read count in peaks

Recall that this version of `R` on Xanadu has many of our libraries
preinstalled. 

```{r engine='bash', eval=F, echo=TRUE}
module load R/4.1.2

R
```


## things to do for the ZNF143

Using the parallel factor ChIP analysis does not work well because
ZNF143 is ablated on chromatin upon degradation. So size factors
cannot be accurately estimated between ZNF143 ChIP degron and
ZNF143/CTCF parallel factor ChIP degron. What is the point of the PF
then? well, we can see that by our peak calling threshold, that all
peaks decrease intensity and are likely true positives when we use
read depth to normalize. 

## Counting reads in peaks and differential binding 

```{r, engine='R', eval=F, echo=T}
#libraries
library(DESeq2)
library(lattice)
library(dplyr)
library(ggplot2)
library(limma)
library(bigWig)

#functions on github
source('https://raw.githubusercontent.com/mjg54/znf143_pro_seq_analysis/master/docs/ZNF143_functions.R')

#new functions
get.counts.interval <- function(df, path.to.bigWig, file.prefix = 'H') {
    vec.names = c()
    inten.df=data.frame(matrix(ncol = 0, nrow = nrow(df)))
    
    for (mod.bigWig in Sys.glob(file.path(path.to.bigWig, paste(file.prefix, "*.bigWig", sep ='')))) {
        factor.name = strsplit(strsplit(mod.bigWig, "/")[[1]][length(strsplit(mod.bigWig, "/")[[1]])], '.bigWig')[[1]][1]
        print(factor.name)
        vec.names = c(vec.names, factor.name)
        loaded.bw = load.bigWig(mod.bigWig)
        print(mod.bigWig)
        mod.inten = bed.region.bpQuery.bigWig(loaded.bw, df, abs.value = TRUE)
        inten.df = cbind(inten.df, mod.inten)
    }
    colnames(inten.df) = vec.names
    r.names = paste(df[,1], ':', df[,2], '-', df[,3], sep='')
    row.names(inten.df) = r.names
    return(inten.df)
}

ma.plot.lattice <- function(ma.df, filename = 'file.name', 
         title.main = "Differential ATAC-seq Accessibility", ymin = -4,
         ymax=4,
         col = c("grey90",  "grey60", "#ce228e" , "#2290cf"))
  {
  pdf(paste("MA_plot_", filename, ".pdf", sep=''), 
      useDingbats = FALSE, width=3.83, height=3.83);
  print(xyplot(ma.df$log2FoldChange ~ log(ma.df$baseMean, base=10),
               groups=ma.df$response,
               col= col, ylim=c(ymin,ymax),
                main=title.main, scales="free", aspect=1, pch=20, cex=0.5,
               ylab=expression("log"[2]~"ChIP-seq change"), 
               xlab=expression("log"[10]~"Mean of Normalized Counts"),
               par.settings=list(par.xlab.text=list(cex=1.1,font=2), 
                                 par.ylab.text=list(cex=1.1,font=2))));
  dev.off()
  }





categorize.deseq.df.repressed <- function(df, fdr = 0.1, log2fold = 0.0, treat
= 'Auxin') {

     df.activated = data.frame(matrix(nrow = 0, ncol = 0))
     df.repressed = data.frame(matrix(nrow = 0, ncol = 0))
	 df.unchanged = data.frame(matrix(nrow = 0, ncol = 0))
     df.dregs = data.frame(matrix(nrow = 0, ncol = 0))
     if (nrow(df[df$padj < fdr & !is.na(df$padj) & df$log2FoldChange > log2fold,]) != 0) {
     	df.activated = df[df$padj < fdr & !is.na(df$padj) & df$log2FoldChange > log2fold,]
	df.activated$response = paste(treat, 'Activated')
	}

     if (nrow(df[df$padj < fdr & !is.na(df$padj) & df$log2FoldChange < -log2fold,]) != 0) {
     	df.repressed = df[df$padj < fdr & !is.na(df$padj) & df$log2FoldChange < -log2fold,]
	df.repressed$response = paste(treat, 'Repressed')
	}
    
    if (nrow(df[df$padj > 0.5 & !is.na(df$padj) & abs(df$log2FoldChange) < 0.25,]) != 0) {
	df.unchanged = df[df$padj > 0.5 & !is.na(df$padj) & abs(df$log2FoldChange) < 0.25,]
    df.unchanged$response = paste(treat, 'Unchanged')
	}

    if (nrow(df[!(df$padj < fdr & !is.na(df$padj) & df$log2FoldChange > log2fold) &
                  !(df$padj < fdr & !is.na(df$padj) & df$log2FoldChange < -log2fold) &
                  !(df$padj > 0.5 & !is.na(df$padj) &
    		  abs(df$log2FoldChange) < 0.25), ]) != 0) {
	df.dregs = df[!(df$padj < fdr & !is.na(df$padj) & df$log2FoldChange > log2fold) &
                  !(df$padj < fdr & !is.na(df$padj) & df$log2FoldChange < -log2fold) &
                  !(df$padj > 0.5 & !is.na(df$padj) &
    		  abs(df$log2FoldChange) < 0.25), ]
	df.dregs$response = paste(treat, 'All Other Genes')
	}
	print(head(df.repressed))
	print(head(df.activated))
	print(head(df.dregs))
	print(head(df.unchanged))
    df.effects.lattice = df.repressed
    #rbind(df.activated, 
    #      df.unchanged, 
    #      df.repressed, 
    #      df.dregs)
	print(head(df.effects.lattice))
    #df.effects.lattice$response = factor(df.effects.lattice$response)
	#df.effects.lattice$response = relevel(df.effects.lattice$response, ref = paste(treat, 'Unchanged'))
	#df.effects.lattice$response = relevel(df.effects.lattice$response, ref = paste(treat, 'All Other Genes'))
    return(df.effects.lattice)
}


#things to do:
# 



#ZNF143 size factors from read depth


x = read.table('CTCF_ChIP_summit_400window_isolated.bed', sep = "\t", header=FALSE)
y = read.table('ZNF143_ChIP_summit_400window.bed', sep = "\t", header=FALSE)

#normalized signal
znf.signal.df = get.counts.interval(y, "/Users/guertinlab/Desktop/ZNF143_ChIP/normalized_bigWig")

#likewise for CTCF
ctcf.signal.df = get.counts.interval(x, "/Users/guertinlab/Desktop/ZNF143_ChIP/normalized_bigWig")


ctcf.counts.df = get.counts.interval(x, "/Users/guertinlab/Desktop/ZNF143_ChIP")
znf143.counts.df = get.counts.interval(y, "/Users/guertinlab/Desktop/ZNF143_ChIP")

#

HA.SF <- read.table("HA_reads.txt", sep = '\t', header = TRUE)[,-1]

HA.size.factors = estimateSizeFactorsForMatrix(HA.SF)



#first step is get sizefactors for CTCFHA degron, CTCFHA ctrl, CTCF degron, CTCF ctrl
#next step is identify ZNF143 peaks distal to any CTCF peak
#get sizefactors of HA ctrl and HACTCF ctrl for these peaks
#get sizefactors of HA degron and HACTCF degron for these peaks 
#use the HACTCF degron and HACTCF ctrl to get normalized size factors between all conditions 

all.regions = rbind(ctcf.counts.df, znf143.counts.df)

#ctcf.all.regions=all.regions[,grepl("CTCF",colnames(all.regions))]
#sample.conditions = factor(sapply(strsplit(colnames(ctcf.all.regions), '_rep'), '[', 1))

#deseq.counts.table = DESeqDataSetFromMatrix(countData = ctcf.all.regions,
#                colData = as.data.frame(sample.conditions), 
#                design = ~ sample.conditions)
				
#dds <- estimateSizeFactors(deseq.counts.table , controlGenes=rownames(deseq.counts.table) %in% rownames(ctcf.counts.df))
#dds <- DESeq(dds)
#anchor.SF = dds$sizeFactor


#
znf143.analysis.regions=all.regions[,grepl("_HA_",colnames(all.regions))]

znf143.analysis.regions = znf143.analysis.regions[rownames(znf143.analysis.regions) %in% rownames(znf143.counts.df),]


znf143.analysis.SF = HA.size.factors

sample.conditions = factor(sapply(strsplit(colnames(znf143.analysis.regions), '_rep'), '[', 1))

deseq.counts.table.4 = DESeqDataSetFromMatrix(countData = znf143.analysis.regions,
                colData = as.data.frame(sample.conditions),
                design = ~ sample.conditions)
		
		
#deseq.counts.table.4$colnames
#HA.sf.final = rep(NA, 6)
#HA.sf = read.table(i, sep = '\t', header = FALSE, check.names=FALSE)[,-1]
#HA.sf=t(HA.sf)

#factors <- read.table("ATAC_sizeFactor.txt")
#HA.sf.final[1:6] = as.numeric(HA.sf[2,1:6])
#names(HA.sf.final) = HA.sf[1,]
#sizeFactors(deseq.df) <- size_factors


#sizeFactors(deseq.counts.table.4) <- HA.sf.final/mean(HA.sf.final)

sizeFactors(deseq.counts.table.4) <- znf143.analysis.SF

dds.4 <- DESeq(deseq.counts.table.4)



DE.results = results(dds.4)
#DE.results.lattice = 
#    categorize.deseq.df.repressed(DE.results, 
#                        fdr = 0.1, log2fold = 0.0, treat =
#						'ZNF143_degron_30min')
						
#DE.results.lattice = 
#    categorize.deseq.df(DE.results, 
#                        fdr = 0.1, log2fold = 0.0, treat =
#						'ZNF143_degron_30min')
						
						
normalized.counts.znf143 = counts(dds.4, normalized=TRUE)
peak.intensities = rowMeans(normalized.counts.znf143[,1:3])

names(peak.intensities) = rownames(normalized.counts.znf143)
chr = sapply(strsplit(names(peak.intensities), ":"), "[", 1)
rnge = sapply(strsplit(names(peak.intensities), ":"), "[", 2)
start =	as.numeric(sapply(strsplit(rnge, "-"), "[", 1)) + 100
end = as.numeric(sapply(strsplit(rnge, "-"), "[", 2)) - 100

quantile(peak.intensities, probs = seq(.05, 1.00, by = .05))

j =0 
q=seq(.05, 1.00, by = .05)
count=0
for (i in quantile(peak.intensities, probs = seq(.05, 1.00, by =
.05)))
{
count = count +1

write.table(file = paste0('quantile', as.character(q[count]), '.bed'), data.frame(chr[peak.intensities > j & peak.intensities <= i], start[peak.intensities > j & peak.intensities <= i], end[peak.intensities > j & peak.intensities <= i], peak.intensities[peak.intensities > j & peak.intensities <= i]), sep = '\t', quote=FALSE, col.names=FALSE, row.names=FALSE )
j = i
}

```


# Now switch over to the PRO-seq vignette to determine the functional peaks

```{r, engine='R', eval=F, echo=T}


length(peak.intensities[peak.intensities >= quantile(peak.intensities, probs = seq(.05, 1.00, by = .05))[13]])

func.znf143.peaks = peak.intensities[peak.intensities >= quantile(peak.intensities, probs = seq(.05, 1.00, by = .05))[13]]

other.znf143.peaks = peak.intensities[peak.intensities < quantile(peak.intensities, probs = seq(.05, 1.00, by = .05))[13]]

```


# motif analysis of other and functional ZNF143 peaks


```{r, engine='R', eval=F, echo=T}

chr = sapply(strsplit(names(func.znf143.peaks), ":"), "[", 1)
rnge = sapply(strsplit(names(func.znf143.peaks), ":"), "[", 2)
start =	as.numeric(sapply(strsplit(rnge, "-"), "[", 1)) + 200
end = as.numeric(sapply(strsplit(rnge, "-"), "[", 2)) - 200

write.table(cbind(chr, start, end), file = "ZNF143_functional_peaks.bed", quote = FALSE,
col.names =FALSE, row.names=FALSE, sep = "\t")


chr = sapply(strsplit(names(other.znf143.peaks), ":"), "[", 1)
rnge = sapply(strsplit(names(other.znf143.peaks), ":"), "[", 2)
start =	as.numeric(sapply(strsplit(rnge, "-"), "[", 1)) + 200
end = as.numeric(sapply(strsplit(rnge, "-"), "[", 2)) - 200

write.table(cbind(chr, start, end), file = "ZNF143_other_peaks.bed", quote = FALSE,
col.names =FALSE, row.names=FALSE, sep = "\t")

```


# deeptools heatmap of ZNF143 functional and non-functional peaks

```{r, engine='bash', eval=F, echo=T}

module load deeptools/3.5.0
module load bedtools

#Make the matrix (this can be used for a composite profile as well)
computeMatrix reference-point --referencePoint center -b 500 -a 500 -p 2 --missingDataAsZero \
  -R ZNF143_functional_peaks.bed \
  -S HEK_ZNF-dTAG_cont_HA.bigWig \
  HEK_ZNF-dTAG_dTAGV_HA.bigWig \
  -o matrix_HA_ChIP_ZNF143_peaks.gz --outFileSortedRegions ZNF143_peaks_sorted_for_heatmap.bed
  
#Make the heatmap
plotHeatmap -m matrix_HA_ChIP_ZNF143_peaks.gz -out heatmap_HA_ChIP_ZNF143_peaks.pdf --heatmapHeight 7 \
   --regionsLabel "ZNF143 peaks" --xAxisLabel "Distance from summit" \
   --samplesLabel "DMSO" "dTAG" --colorMap Purples -min 0 -max 300 --whatToShow "heatmap and colorbar" 
   
   
   #Make the matrix (this can be used for a composite profile as well)
computeMatrix reference-point --referencePoint center -b 500 -a 500 -p 2 --missingDataAsZero \
  -R ZNF143_other_peaks.bed \
  -S HEK_ZNF-dTAG_cont_HA.bigWig \
  HEK_ZNF-dTAG_dTAGV_HA.bigWig \
  -o all_matrix_HA_ChIP_ZNF143_peaks.gz --outFileSortedRegions other_ZNF143_peaks_sorted_for_heatmap.bed
  
#Make the heatmap
plotHeatmap -m all_matrix_HA_ChIP_ZNF143_peaks.gz -out other_heatmap_HA_ChIP_ZNF143_peaks.pdf --heatmapHeight 7 \
   --regionsLabel "ZNF143 peaks" --xAxisLabel "Distance from summit" \
   --samplesLabel "DMSO" "dTAG" --colorMap Purples -min 0 -max 300 --whatToShow "heatmap and colorbar" 

#so we can see peaks
plotHeatmap -m all_matrix_HA_ChIP_ZNF143_peaks.gz -out other_heatmap_HA_ChIP_ZNF143_peaks_scaled.pdf --heatmapHeight 7 \
   --regionsLabel "ZNF143 peaks" --xAxisLabel "Distance from summit" \
   --samplesLabel "DMSO" "dTAG" --colorMap Purples -min 0 -max 13 --whatToShow "heatmap and colorbar" 
```


# De novo motif analysis of ZNF143 peaks

I am brute forcing my way through an iterative and exhaustive motif
analysis. This should be written into a systematic loop. The reason I
did not is because I am manualy looking at the top hit to ensure that
it is a ZNF143 motif variant before going on to the next
iteration. Any student using this for their work needs to put in the
work to automate systematic iterative exhaustive de novo motif
analysis. 

```{r engine='bash', eval=F, echo=TRUE}
module load meme/5.4.1
module load bedtools

genome=/home/FCAM/meds5420/genomes/hg38.fa
sizes=/home/FCAM/meds5420/genomes/hg38.chrom.sizes

for i in ZNF143_*_peaks.bed
do
	name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_peaks.bed" '{print $1}')
	echo $name
	slopBed -b 70 -i $i -g $sizes  | sort -k1,1 -k2,2n > ${name}_summit_140window.bed
   	slopBed -b 200 -i $i -g $sizes  | sort -k1,1 -k2,2n > ${name}_summit_400window.bed
	fastaFromBed -fi $genome -bed ${name}_summit_140window.bed -fo ${name}_summit_140window.fasta
done

for i in ZNF143_*_peaks.bed
do
	name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_peaks.bed" '{print $1}')
	echo $name
	meme -p 16 -oc ${name}_motif_functional.meme_output -nmotifs 1 -objfun classic -csites 20000 -searchsize 0 -minw 10 -maxw 25 -revcomp -dna -markov_order 3 -maxsize 100000000 ${name}_summit_140window.fasta
done


fimo --max-strand --max-stored-scores 10000000 --oc ZNF143_functional_motif_functional.fimo_output ZNF143_functional_motif_functional.meme_output/meme.txt $genome

tail -n +2 ZNF143_functional_motif_functional.fimo_output/fimo.tsv | awk '{OFS="\t";} {print $3,$4,$5,$7,$8,$6}' | grep -v Individual| grep -v format | grep -v "max-stored-scores" | sort -k1,1 -k2,2n | tail -n +2 > fimo_ZNF143_functional_round1.bed

intersectBed -v -wa -a ZNF143_functional_summit_140window.bed -b fimo_ZNF143_functional_round1.bed > ZNF143_ChIP_summit_140window_no_func_motif_1.bed

fastaFromBed -fi $genome -bed ZNF143_ChIP_summit_140window_no_func_motif_1.bed -fo ZNF143_ChIP_summit_140window_no_func_motif_1.fasta

meme -p 16 -oc ZNF143_functional_motif_functional_second.meme_output -nmotifs 1 -objfun classic -csites 20000 -searchsize 0 -minw 10 -maxw 25 -revcomp -dna -markov_order 3 -maxsize 100000000 ZNF143_ChIP_summit_140window_no_func_motif_1.fasta

fimo --max-strand --max-stored-scores 10000000 --oc ZNF143_functional_motif_functional_second.fimo_output ZNF143_functional_motif_functional_second.meme_output/meme.txt $genome

tail -n +2 ZNF143_functional_motif_functional_second.fimo_output/fimo.tsv | awk '{OFS="\t";} {print $3,$4,$5,$7,$8,$6}' | grep -v Individual| grep -v format | grep -v "max-stored-scores" | sort -k1,1 -k2,2n | tail -n +2 > fimo_ZNF143_functional_round2.bed

intersectBed -v -wa -a ZNF143_ChIP_summit_140window_no_func_motif_1.bed -b fimo_ZNF143_functional_round2.bed > ZNF143_ChIP_summit_140window_no_func_motif_1_or_2.bed

fastaFromBed -fi $genome -bed ZNF143_ChIP_summit_140window_no_func_motif_1_or_2.bed -fo ZNF143_ChIP_summit_140window_no_func_motif_1_or_2.fasta

meme -p 16 -oc ZNF143_functional_motif_functional_third.meme_output -nmotifs 1 -objfun classic -csites 20000 -searchsize 0 -minw 10 -maxw 25 -revcomp -dna -markov_order 3 -maxsize 100000000 ZNF143_ChIP_summit_140window_no_func_motif_1_or_2.fasta

#this returned garbage

wc -l ZNF143_ChIP_summit_140window_no_func_motif_1_or_2.bed
wc -l ZNF143_functional_summit_140window.bed

#96% have either motif
# what comparison comparison set of regions can we use as a control?
```

<!--
# make a barchart for fraction of peaks with motifs 

```{r engine='bash', eval=F, echo=TRUE}
#input for barchart

rm barchart.txt
for i in quantile*.bed
do
	name=$(echo $i | awk -F"quantile" '{print $NF}' | awk -F".bed" '{print $1}')
	sort -k1,1 -k2,2n $i > tmp.bed
	echo $name 
	intersectBed -u -wa -a tmp.bed -b fimo_ZNF143.bed fimo_ZNF143_motif2.bed > overlap_${name}_znf143.bed
	num=`wc -l overlap_${name}_znf143.bed | cut -f1 -d 'o'`
	tot=`wc -l $i | cut -f1 -d 'q'`
	echo ${num} pos ${name} ZNF143 >> barchart.txt
	neg=$(expr ${tot} - ${num})
	echo ${neg} neg ${name} ZNF143 >> barchart.txt
done


```


-->

# Jinhong: To do

## run through the entire vignette and add context paragraphs and include figures

## make a composite PSWM of motif 1 and motif 2 from the functional peaks

## make a composite FIMO motif signal at peak summits

## de novo motif analysis on other (non-functional peaks)

## Calculate the fraction of reads in peaks (FRiP) score for:

### the full set of ZNF143 narrowPeaks

### functional peaks (use narrowPeaks file)

### other (non-func) peaks (use narrowPeaks file)

<!--

```{r engine='R', eval=F, echo=TRUE}


plot.barchart <- function(df.barchart, filename = "~/Desktop/barchart.beds.broadPeaks.pdf") {
  pdf(filename, width=12.83, height=6)

  polycol <- trellis.par.get("superpose.polygon")
  polycol$col <- c("red", "grey40")
  trellis.par.set("superpose.polygon",polycol)


  print(barchart(as.numeric(as.character(fraction))~factor, data = df.barchart, groups=p_n,
                 stack=TRUE,
                 as.table=TRUE,
                 layout=c(1,1),
               #auto.key = list(title = "",rows=3,fill=colors,just="bottom"),
                 main="H3R26Cit Peaks",
               #xlab = "HSF1 or HSF2 Peaks",
                 ylab="Fraction Enriched Red",
                 cex.axis=1.2,
                 between=list(y=0.5, x=0.5),
                 font.axis=1,
                 par.settings=list(par.xlab.text=list(cex=1.0,font=1),
                   par.ylab.text=list(cex=1.0,font=1),
                   axis.text=list(cex=1,font=1),
                   strip.background=list(col="#ecdaf5"),
                   par.main.text=list(cex=1.2, font=1)),
                                        #aspect = 1,
                 scales=list(x=list(alternating=c(1,1,1,0,0,0),rot=45),
                   y=list(alternating=c(1,1)))))
  
  dev.off()
}

```








```{r, engine='R', eval=F, echo=T}

DE.results = results(dds.4)



categorize.deseq.df.repressed <- function(df, fdr = 0.1, log2fold = 0.0, treat
= 'Auxin') {

     df.activated = data.frame(matrix(nrow = 0, ncol = 0))
     df.repressed = data.frame(matrix(nrow = 0, ncol = 0))
	 df.unchanged = data.frame(matrix(nrow = 0, ncol = 0))
     df.dregs = data.frame(matrix(nrow = 0, ncol = 0))
     if (nrow(df[df$padj < fdr & !is.na(df$padj) & df$log2FoldChange > log2fold,]) != 0) {
     	df.activated = df[df$padj < fdr & !is.na(df$padj) & df$log2FoldChange > log2fold,]
	df.activated$response = paste(treat, 'Activated')
	}

     if (nrow(df[df$padj < fdr & !is.na(df$padj) & df$log2FoldChange < -log2fold,]) != 0) {
     	df.repressed = df[df$padj < fdr & !is.na(df$padj) & df$log2FoldChange < -log2fold,]
	df.repressed$response = paste(treat, 'Repressed')
	}
    
    if (nrow(df[df$padj > 0.5 & !is.na(df$padj) & abs(df$log2FoldChange) < 0.25,]) != 0) {
	df.unchanged = df[df$padj > 0.5 & !is.na(df$padj) & abs(df$log2FoldChange) < 0.25,]
    df.unchanged$response = paste(treat, 'Unchanged')
	}

    if (nrow(df[!(df$padj < fdr & !is.na(df$padj) & df$log2FoldChange > log2fold) &
                  !(df$padj < fdr & !is.na(df$padj) & df$log2FoldChange < -log2fold) &
                  !(df$padj > 0.5 & !is.na(df$padj) &
    		  abs(df$log2FoldChange) < 0.25), ]) != 0) {
	df.dregs = df[!(df$padj < fdr & !is.na(df$padj) & df$log2FoldChange > log2fold) &
                  !(df$padj < fdr & !is.na(df$padj) & df$log2FoldChange < -log2fold) &
                  !(df$padj > 0.5 & !is.na(df$padj) &
    		  abs(df$log2FoldChange) < 0.25), ]
	df.dregs$response = paste(treat, 'All Other Genes')
	}
	print(head(df.repressed))
	print(head(df.activated))
	print(head(df.dregs))
	print(head(df.unchanged))
    df.effects.lattice = df.repressed
    #rbind(df.activated, 
    #      df.unchanged, 
    #      df.repressed, 
    #      df.dregs)
	print(head(df.effects.lattice))
    #df.effects.lattice$response = factor(df.effects.lattice$response)
	#df.effects.lattice$response = relevel(df.effects.lattice$response, ref = paste(treat, 'Unchanged'))
	#df.effects.lattice$response = relevel(df.effects.lattice$response, ref = paste(treat, 'All Other Genes'))
    return(df.effects.lattice)
}






DE.results.lattice = 
    categorize.deseq.df(DE.results, 
                        fdr = 0.1, log2fold = 0.0, treat =
						'ZNF143_degron_30min')
						
DE.results.lattice = 
    categorize.deseq.df.repressed(DE.results, 
                        fdr = 0.1, log2fold = 0.0, treat = 'ZNF143_degron_30min')						
						
head(DE.results.lattice)
repressed.all = DE.results.lattice[DE.results.lattice$response == 
                                             'ZNF143_degron_30min Repressed',]
dim(repressed.all)

activated.all = DE.results.lattice[DE.results.lattice$response == 
                                             'ZNF143_degron_30min Activated',]
dim(activated.all)
# what is happening in these 213 regions? likely no ZNF143 motif

ma.plot.lattice(DE.results.lattice, filename = 'ZNF143_30min_degradation_SF', 
        title.main = "Differential Binding", ymin=-12, ymax=4)		
		

DE.results = results(dds.4)
# also get un normalized ZNF143 MA plot.

#NEXT FOR EACH QUANTILE PLOT THE FRACTION THAT SIGNIFICANTLY DECREASE INTENSITY

#goal is to use a combination of peaks with motifs in each quantile
and sensitivity to depleteion to get a peak cut off.
STOPPED HERE


#make a MA plot without normalized SF to show theimportance of PF chip
# de novo motif analysi sof the NOT ACTIVATED to see if the ZNF143
#motif comes out. If so we are being too conservative in the
#differntial binding analysis and recommend moving to aligned Reads as
#size factors

DE.results = results(dds)

head(DE.results)

DE.results.lattice = 
    categorize.deseq.df(DE.results, 
                        fdr = 0.1, log2fold = 0.0, treat = 'ZNF143_degron_30min')

head(DE.results.lattice)


repressed.all = DE.results.lattice[DE.results.lattice$response == 
                                             'ZNF143_degron_30min Repressed',]

dim(repressed.all)


ma.plot.lattice(DE.results.lattice, filename = 'ZNF143_30min_degradation', 
        title.main = "Differential Binding")		
		
#we are not ranking by fold change, so no need for shrinkage

#lets export a smaller genomic interval for each region for subsequent motif analysis

chr = sapply(strsplit(rownames(activated.all), ":"), "[", 1)
rnge = sapply(strsplit(rownames(activated.all), ":"), "[", 2)
start =	as.numeric(sapply(strsplit(rnge, "-"), "[", 1)) + 100
end = as.numeric(sapply(strsplit(rnge, "-"), "[", 2)) - 100

write.table(cbind(chr, start, end), file = "ATAC_increase_TRPS1_degron.bed", quote = FALSE,
col.names =FALSE, row.names=FALSE, sep = "\t")


unchanged.all = DE.results.lattice[DE.results.lattice$response == 
                                             'TRPS1_degron_30min Unchanged',]
											 
chr.un = sapply(strsplit(rownames(unchanged.all), ":"), "[", 1)
rnge.un = sapply(strsplit(rownames(unchanged.all), ":"), "[", 2)
start.un =	as.numeric(sapply(strsplit(rnge.un, "-"), "[", 1)) + 100
end.un = as.numeric(sapply(strsplit(rnge.un, "-"), "[", 2)) - 100

write.table(cbind(chr.un, start.un, end.un), file = "ATAC_unchanged_TRPS1_degron.bed", quote = FALSE,
col.names =FALSE, row.names=FALSE, sep = "\t")



#motifs
with.motif = read.table('ZNF143_ChIP_summit_140window_w_motifs.bed', sep = "\t", header=FALSE)
no.motif = read.table('ZNF143_ChIP_summit_140window_no_motifs.bed', sep = "\t", header=FALSE)
with.motif = cbind(with.motif, 'ZNF143 Motif')
no.motif = cbind(no.motif, 'No Motif')

colnames(with.motif) = c('chr', 'start', 'end', 'peak', 'strength', 'category')
colnames(no.motif) = c('chr', 'start', 'end', 'peak', 'strength', 'category')

motif.strength = rbind(with.motif, no.motif)

densityplot(~log(strength, base = 10), data=motif.strength,
	groups = category,
       xlab="Peak Intensity",
        scales="free",
		par.settings = list(superpose.line = list(col = c("#5304a0","#0e930a"))),
   aspect=1, lwd =3,
       plot.points=FALSE,auto.key=TRUE,
	   panel=function(x,...){
      panel.densityplot(x,...)
      panel.abline(v=1, col.line="red", lty = 2) 
    })
#remove those with intensities less than 10 (no motif)

#data processing and plotting
#just calculate sizeFactors from read depth
#use aligned read depth


```

 don't use below 0.7 quantile

62 fold dynamic range, there is littel evidence of PRO changes near
genes below 1.6% of the max peak height








# DID NOT DO THIS YET De novo motif analysis of differential peaks

Lastly, we perform _de novo_ motif analysis as we have previously with
`meme`. We can also determine the fraction of regions with immediate
increases in accessibility have the underlying motif. However, this
value is meaningless without a comparison. We could compare to
`TRPS1_degron_30min Unchanged` region set.

```{r engine='bash', eval=F, echo=TRUE}
module load bedtools
#NEXT 
#do meme on those without a canonical ZNF143

rm barchart.txt
for i in quantile*.bed
do
	name=$(echo $i | awk -F"quantile" '{print $NF}' | awk -F".bed" '{print $1}')
	sort -k1,1 -k2,2n $i > tmp.bed
	echo $name 
	intersectBed -u -wa -a tmp.bed -b fimo_ZNF143.bed fimo_ZNF143_motif2.bed > overlap_${name}_znf143.bed
	num=`wc -l overlap_${name}_znf143.bed | cut -f1 -d 'o'`
	tot=`wc -l $i | cut -f1 -d 'q'`
	echo ${num} pos ${name} ZNF143 >> barchart.txt
	neg=$(expr ${tot} - ${num})
	echo ${neg} neg ${name} ZNF143 >> barchart.txt
done

num	p_n	b_f	factor	fraction
66	pos	bottom	BEAF_70	  .04737975
0	neg	bottom	BEAF_70	  0
1327	neu	bottom	BEAF_70	  .95262025

```




```{r engine='bash', eval=F, echo=TRUE}
genome=/home/FCAM/meds5420/genomes/hg38.fa
module load bedtools

for i in quantile*.bed
do
	name=$(echo $i | awk -F"/" '{print $NF}' | awk -F".bed" '{print $1}')
	echo $name
	fastaFromBed -fi $genome -bed $i -fo $name.fasta
done

DO MAST IN THESE WITH THE TOP DE NOVO MOTIF.

```


-->

<!-- 
```{r engine='bash', eval=F, echo=TRUE}

#Defined functional TRPS1 peaks by MACS2 q-value < 10^(-30)
awk '$5 > 30' /scratch/ts2hx/ChIP/results/macs2/HA_DMSO_vs_dTAG_summits.bed > \
  Functional_TRPS1_summits.bed

#Make the matrix (this can be used for a composite profile as well)
computeMatrix reference-point --referencePoint center -b 500 -a 500 -p 20 --missingDataAsZero \
  -R ZNF143_ChIP_summits_final.bed \
  -S HEK_ZNF-dTAG_cont_HA.bigWig \
  HEK_ZNF-dTAG_dTAGV_HA.bigWig \
  -o matrix_HA_ChIP_ZNF143_peaks.gz --outFileSortedRegions ZNF143_peaks_sorted_for_heatmap.bed
  
#Make the heatmap
plotHeatmap -m matrix_HA_ChIP_ZNF143_peaks.gz -out heatmap_HA_ChIP_ZNF143_peaks.pdf --heatmapHeight 14 \
   --regionsLabel "ZNF143 peaks" --xAxisLabel "Distance from summit" \
   --samplesLabel "DMSO" "dTAG" --colorMap Purples --whatToShow "heatmap and colorbar" 

```

Now we want to identify all CTCF peaks that are distal to ZNF143
binding sites and identify all ZNF143 peaks.

```{r engine='bash', eval=F, echo=TRUE}
module load bedtools
intersectBed -wa -v -a ZNF143_ChIP_peaks.narrowPeak -b CTCF_ChIP_peaks.narrowPeak > ZNF_away_CTCF.narrowPeak
intersectBed -wa -v -a CTCF_ChIP_peaks.narrowPeak -b ZNF143_ChIP_peaks.narrowPeak > CTCF_away_ZNF.narrowPeak
intersectBed -wa -a CTCF_ChIP_summit_400window.bed -b CTCF_away_ZNF.narrowPeak | sort -k1,1 -k2,2n | uniq > CTCF_ChIP_summit_400window_isolated.bed


closestBed -d -a ZNF143_ChIP_summit_400window.bed -b CTCF_ChIP_summits_final.bed | awk '{OFS="\t";} {print $1, $2, $3, $11}' > ZNF143_CTCFdistance.bed

genome=/home/FCAM/meds5420/genomes/hg38.fa
fastaFromBed -fi $genome -bed ZNF143_ChIP_summit_140window.bed -fo ZNF143_ChIP_summit_140window.fasta

# inputs to bigWig in R
head CTCF_ChIP_summit_400window_isolated.bed
head ZNF143_ChIP_summit_400window.bed


#size factors

module load samtools/1.12
for i in HA
do
  echo $i
  > ${i}_header.txt
  > ${i}_reads.txt
  for j in HEK_ZNF-dTAG*_${i}_*.bam
  do
    echo $j
    name=$(echo $j | awk -F".sorted.bam" '{print $1}')
    echo $name | paste ${i}_header.txt - > ${i}_tmp.txt 
    mv ${i}_tmp.txt ${i}_header.txt
    reads=`samtools view -c -f 0x3 $j`
    echo $reads | paste ${i}_reads.txt - > ${i}_tmp.txt 
    mv ${i}_tmp.txt ${i}_reads.txt
  done  
  cat ${i}_header.txt ${i}_reads.txt > ${i}_tmp.txt
  mv ${i}_tmp.txt ${i}_reads.txt
  rm ${i}_header.txt
done 

```



# De novo motif analysis of ZNF143 peaks

I am brute forcing my way through an iterative and exhaustive motif
analysis. I need to write this into a systematic loop. The reason I
did not is because I am manualy looking at the top hit to ensure that
it is a ZNF143 motif variant before going on to the next iteration.

```{r engine='bash', eval=F, echo=TRUE}
module load meme/5.4.1
genome=/home/FCAM/meds5420/genomes/hg38.fa

meme -p 16 -oc ZNF143_motif1.meme_output -nmotifs 3 -objfun classic -csites 20000 -searchsize 0 -minw 10 -maxw 25 -revcomp -dna -markov_order 3 -maxsize 100000000 ZNF143_ChIP_summit_140window.fasta

fimo --motif CTGGGAAATGTAGT --max-strand --max-stored-scores 10000000 --oc fimo_ZNF143 ZNF143_csites.meme_output/meme.txt $genome

tail -n +2 fimo.tsv | awk '{OFS="\t";} {print $3,$4,$5,$7,$8,$6}' | grep -v Individual| grep -v CTGGGAAATGTAGT | grep -v format | sort -k1,1 -k2,2n | tail -n +2 > fimo_ZNF143.bed
 
intersectBed -v -wa -a ZNF143_ChIP_summit_140window.bed -b fimo_ZNF143.bed > ZNF143_ChIP_summit_140window_no_motif1.bed
 
fastaFromBed -fi $genome -bed ZNF143_ChIP_summit_140window_no_motif1.bed -fo ZNF143_ChIP_summit_140window_no_motif1.fasta
 
meme -p 16 -oc ZNF143_motif2.meme_output -nmotifs 3 -objfun classic -csites 20000 -searchsize 0 -minw 10 -maxw 25 -revcomp -dna -markov_order 3 -maxsize 100000000 ZNF143_ChIP_summit_140window_no_motif1.fasta

fimo --thresh 0.001 --motif YTCCCAGMATSCHYYGCG --max-strand --max-stored-scores 10000000 --oc fimo_ZNF143_motif1 ZNF143_motif2.meme_output/meme.txt $genome

tail -n +2 fimo_ZNF143_motif1/fimo.tsv | awk '{OFS="\t";} {print $2,$3,$4,$6,$7,$5}' | grep -v Individual| grep -v YTCCCAGMATSCHYYGCG | grep -v format | sort -k1,1 -k2,2n | tail -n +2 > fimo_ZNF143_motif2.bed
 
intersectBed -v -wa -a ZNF143_ChIP_summit_140window.bed -b fimo_ZNF143.bed fimo_ZNF143_motif2.bed > ZNF143_ChIP_summit_140window_no_motifs.bed

fastaFromBed -fi $genome -bed ZNF143_ChIP_summit_140window_no_motifs.bed -fo ZNF143_ChIP_summit_140window_no_motifs.fasta
 
meme -p 16 -oc ZNF143_motif3.meme_output -nmotifs 3 -objfun classic -csites 20000 -searchsize 0 -minw 10 -maxw 25 -revcomp -dna
-markov_order 3 -maxsize 100000000 ZNF143_ChIP_summit_140window_no_motifs.fasta

fimo --thresh 0.0001 --motif SMVVVNNVRRRACTACATTTCCCAG --max-strand --max-stored-scores 10000000 --oc fimo_ZNF143_motif3 ZNF143_motif3.meme_output/meme.txt $genome

#stopped here

tail -n +2 fimo_ZNF143_motif3/fimo.tsv | awk '{OFS="\t";} {print $3,$4,$5,$7,$8,$6}' | grep -v Individual| grep -v SMVVVNNVRRRACTACATTTCCCAG | grep -v format | sort -k1,1 -k2,2n | tail -n +2 > fimo_ZNF143_motif3.bed


#the above should be in some sort of loop, but at least rechecked, because I hacked it together

intersectBed -v -a ZNF143_ChIP_summit_140window.bed -b fimo_ZNF143.bed fimo_ZNF143_motif2.bed fimo_ZNF143_motif3.bed > ZNF143_ChIP_summit_140window_no_motifs.bed

intersectBed -wa -u -a ZNF143_ChIP_summit_140window.bed -b fimo_ZNF143.bed fimo_ZNF143_motif2.bed fimo_ZNF143_motif3.bed > ZNF143_ChIP_summit_140window_w_motifs.bed
```



x = read.table('CTCF_ChIP_summit_400window_isolated.bed', sep = "\t", header=FALSE)
y = read.table('ZNF143_ChIP_summit_400window.bed', sep = "\t", header=FALSE)

x = x[((ctcf.signal.df[,'HEK_ZNF-dTAG_cont_CTCF']/ctcf.signal.df[,'HEK_ZNF-dTAG_cont_IgG'])>10) & ctcf.signal.df[,'HEK_ZNF-dTAG_cont_CTCF'] > max(ctcf.signal.df[,'HEK_ZNF-dTAG_cont_CTCF']/200),]

z = read.table('ZNF143_CTCFdistance.bed', sep = "\t", header=FALSE)

z.regions=paste0(z[,1],z[,2])
y.regions=paste0(y[,1],y[,2])

z =z[z.regions %in% y.regions,]
z.regions=paste0(z[,1],z[,2])
z =z[!duplicated(z.regions),]

distal.znf143 = z[z[,4]<2000,]


intervals.znf = paste(distal.znf143[,1], ':', distal.znf143[,2], '-', distal.znf143[,3], sep='')
distal.znf143 = distal.znf143[!duplicated(intervals.znf),]
intervals.znf = intervals.znf[!duplicated(intervals.znf)]




DE.results.1 = results(dds, contrast=c("sample.conditions","HEK_ZNF-dTAG_cont_CTCF","HEK_ZNF-dTAG_cont_HACTCF")) 

DE.results.lattice.1 = 
    categorize.deseq.df(DE.results.1, 
                        fdr = 0.1, log2fold = 0.0, treat = 'ZNF143_degron_30min_CTCF')
ma.plot.lattice(DE.results.lattice.1, filename = 'ZNF143_30min_degradation_CTCFpeaks_SFcont_CTCF_vs_HACTCF', 
        title.main = "Differential Binding", ymin=-6, ymax=6)	

#is every peak repressed on this near a ZNF143 binding site that we did not call?
#motif analysis

DE.results.11 = results(dds, contrast=c("sample.conditions","HEK_ZNF-dTAG_cont_CTCF","HEK_ZNF-dTAG_dTAGV_CTCF")) 

DE.results.lattice.11 = 
    categorize.deseq.df(DE.results.11, 
                        fdr = 0.1, log2fold = 0.0, treat = 'ZNF143_degron_30min_CTCF')
ma.plot.lattice(DE.results.lattice.11, filename = 'ZNF143_30min_degradation_CTCFpeaks_SF_CTCF_ctrl_degron', 
        title.main = "Differential Binding", ymin=-6, ymax=6)	


DE.results.12 = results(dds, contrast=c("sample.conditions","HEK_ZNF-dTAG_dTAGV_CTCF","HEK_ZNF-dTAG_dTAGV_HACTCF")) 

DE.results.lattice.12 = 
    categorize.deseq.df(DE.results.12, 
                        fdr = 0.1, log2fold = 0.0, treat = 'ZNF143_degron_30min_CTCF')
ma.plot.lattice(DE.results.lattice.12, filename = 'ZNF143_30min_degradation_CTCFpeaks_SF_CTCF_pf_degron', 
        title.main = "Differential Binding", ymin=-6, ymax=6)	



DE.results.13 = results(dds, contrast=c("sample.conditions","HEK_ZNF-dTAG_dTAGV_HACTCF","HEK_ZNF-dTAG_dTAGV_HACTCF")) 

DE.results.lattice.13 = 
    categorize.deseq.df(DE.results.13, 
                        fdr = 0.1, log2fold = 0.0, treat = 'ZNF143_degron_30min_CTCF')
ma.plot.lattice(DE.results.lattice.13, filename = 'ZNF143_30min_degradation_CTCFpeaks_SF_upon_degron_CTCF_vs_CTCF', 
        title.main = "Differential Binding", ymin=-6, ymax=6)	



#this needs to be filtered for ZNF143 peaks beign considered.
#???--this is ALL FUCKED UP
#control HA SF

#FEM1A is a good pfChIP 

ctrl.znf143 = c("chr5:173616547-173616948", "chr12:6852316-6852717", "chr1:156767180-156767581")

contHA.all.regions=all.regions[,grepl("cont_HA",colnames(all.regions))]
sample.conditions = factor(sapply(strsplit(colnames(contHA.all.regions), '_rep'), '[', 1))
deseq.counts.table.2 = DESeqDataSetFromMatrix(countData = contHA.all.regions,
                colData = as.data.frame(sample.conditions), 
                design = ~ sample.conditions)

#dds.2 <- estimateSizeFactors(deseq.counts.table.2, controlGenes=rownames(deseq.counts.table.2) %in% intervals.znf)
dds.2 <- estimateSizeFactors(deseq.counts.table.2, controlGenes=rownames(deseq.counts.table.2) %in% ctrl.znf143)
dds.2 <- DESeq(dds.2)

znfpeaks.SF = dds.2$sizeFactor

DE.results.2 = results(dds.2)
DE.results.lattice.2 = 
    categorize.deseq.df(DE.results.2, 
                        fdr = 0.1, log2fold = 0.0, treat = 'ZNF143_degron_30min')
ma.plot.lattice(DE.results.lattice.2, filename = 'ZNF143_30min_degradation_ZNFpeaks_SF', 
        title.main = "Differential Binding", ymin=-6, ymax=6)		
		
#it is clear from this MA plot that the ZNF143 overlap CTCF somewhat.




#degron HA SF
dtagHA.all.regions=all.regions[,grepl("dTAGV_HA",colnames(all.regions))]
sample.conditions = factor(sapply(strsplit(colnames(dtagHA.all.regions), '_rep'), '[', 1))
deseq.counts.table.3 = DESeqDataSetFromMatrix(countData = dtagHA.all.regions,
                colData = as.data.frame(sample.conditions), 
                design = ~ sample.conditions)
				

#dds.3 <- estimateSizeFactors(deseq.counts.table.3, controlGenes=rownames(deseq.counts.table.3) %in% intervals.znf)
dds.3 <- estimateSizeFactors(deseq.counts.table.3, controlGenes=rownames(deseq.counts.table.3) %in% ctrl.znf143)
dds.3 <- DESeq(dds.3)


DE.results.3 = results(dds.3)
DE.results.lattice.3 = 
    categorize.deseq.df(DE.results.3, 
                        fdr = 0.1, log2fold = 0.0, treat = 'ZNF143_degron_30min')
ma.plot.lattice(DE.results.lattice.3, filename = 'ZNF143_30min_degradation_ZNFpeaks_degron_SF', 
        title.main = "Differential Binding", ymin=-8, ymax=8)		
		




degronpeaks.SF = dds.3$sizeFactor
names(degronpeaks.SF)
names(znfpeaks.SF)
names(anchor.SF)

anchor.SF[names(anchor.SF) %in% names(degronpeaks.SF)]
degronpeaks.SF[names(degronpeaks.SF) %in% names(anchor.SF)]
mean.anchor.SF = mean(anchor.SF[names(anchor.SF) %in% names(degronpeaks.SF)])
mean.degron.SF = mean(degronpeaks.SF[names(degronpeaks.SF) %in% names(anchor.SF)])

norm.degronpeaks.SF = degronpeaks.SF *(mean.anchor.SF/mean.degron.SF)



anchor.SF[names(anchor.SF) %in% names(znfpeaks.SF)]
znfpeaks.SF[names(znfpeaks.SF) %in% names(anchor.SF)]
mean.anchor.SF = mean(anchor.SF[names(anchor.SF) %in% names(znfpeaks.SF)])
mean.degron.SF = mean(znfpeaks.SF[names(znfpeaks.SF) %in% names(anchor.SF)])

norm.znfpeaks.SF = znfpeaks.SF *(mean.anchor.SF/mean.degron.SF)

#then maybe we average the common SF and incorporate these new ones and compare to read depth SF?

all.SF = rep(NA, 21)
all.SF[1:6] = anchor.SF[1:6]
all.SF[4:6] = (anchor.SF[4:6]+norm.znfpeaks.SF[1:3])/2
all.SF[7:9] = norm.znfpeaks.SF[4:6]
#10:12 are IgG
all.SF[13:15] = anchor.SF[7:9]
all.SF[16:18] = (anchor.SF[10:12]+norm.degronpeaks.SF[1:3])/2
all.SF[19:21] = norm.degronpeaks.SF[4:6]
names(all.SF) = colnames(all.regions)

-->
