---
title: "GATA3_ATAC_analysis"

author:
- Siyu Sun
- Michael J. Guertin
header-includes:
- \usepackage{color}
- \usepackage{float}
- \DeclareUnicodeCharacter{2212}{-}
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    toc: true
fontsize: 14pt
geometry: margin=1in
---


# Install required packages
Install `bigWig` and `latticeExtra` package \
```{r engine='R', eval=F, echo=T}
install.packages("devtools", quiet = TRUE)
library(devtools)
devtools::install_github('andrelmartins/bigWig',
              subdir='bigWig')
library(bigWig)

install.packages("DESeq2", quiet = TRUE)
install.packages("dplyr", quiet = TRUE)
```

Install `bedtools` \
```{r engine='bash', eval=F, echo=T}
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
brew install bedtools
```

Install `Biostrings` \
```{r engine='R', eval=F, echo=T}
if (!requireNamespace("Biostrings", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
  BiocManager::install("Biostrings")
}
```

Install `latticeExtra` \
```{r engine='R', eval=TRUE, echo=T}
install.packages("latticeExtra", quiet = TRUE)
```

# Renaming
```{r engine='bash', eval=TRUE, echo=T}
for i in *R1_001.fastq.gz
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_S" '{print $1}')
    cat ${name}*R1_001.fastq.gz > MCF7_dTAGGATAClone522_30min_${name}_PE1.fastq.gz
    cat ${name}*R2_001.fastq.gz > MCF7_dTAGGATAClone522_30min_${name}_PE2.fastq.gz    
done
```

# Cut off the adapter with `cutadapt`

`-j`: the number of cores; \
`-m` specifies the minimal length of a read to keep after adapter sequence removal; \
`-O` is the number of bases to trim off the end of the read if it overlaps with the adapter sequence. \
`-a` and `-A` options are the adapter sequences of the PE1 and PE2 reads. (Tn5 mosaic end sequence: CTGTCTCTTATACACATCT). \
`-o` (PE1) and `-p` (PE2) specfies the output file. \
The last two positional areguments are the input fastq files. We save the output to a log file. \

```{r engine='bash', eval=TRUE, echo=T}
#! /bin/sh
#SBATCH --job-name=GATA_atac_cutadapt.sh    
#SBATCH -N 1
#SBATCH -n 1
#SBATCH -c 8
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=32G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o cutadapt.sh_%j.out
#SBATCH -e cutadapt.sh_%j.err

hostname
module load cutadapt/3.5

for i in ../raw_fastq_renamed/*PE1.fastq
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1" '{print $1}')
    echo $name
    echo unzipping $i
    #gunzip $i
    echo unzipping ${name}_PE2.fastq.gz
    gunzip ../raw_fastq_renamed/${name}_PE2.fastq.gz
    cutadapt -a CTGTCTCTTATACACATCT -A CTGTCTCTTATACACATCT -j 8 -m 10 -O 1 -o ${name}_PE1_no_adapt.fastq -p ${name}_PE2_no_adapt.fastq ../raw_fastq_renamed/${name}_PE1.fastq ../raw_fastq_renamed/${name}_PE2.fastq 2>&1 | tee ${name}_cutadapt.log

done
```


# Alighment

## Filter reads that align to the mitochondrial chromosome
```{r engine='bash', eval=TRUE, echo=T}
#! /bin/sh
#SBATCH --job-name=humanchrM_bt2.sh     
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 4                 
#SBATCH -p general           
#SBATCH --qos=general     
#SBATCH --mem=10G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o humanchrM_bt2.sh_%j.out
#SBATCH -e humanchrM_bt2.sh_%j.err

module load genometools/1.5.10
module load bowtie2
export PATH=$PATH:/home/FCAM/ssun/packages

# build the mitochondrial genome
wget https://hgdownload.cse.ucsc.edu/goldenpath/hg38/chromosomes/chrM.fa.gz
gunzip chrM.fa.gz
bowtie2-build chrM.fa chrM
```

Since mitochondrial DNA is preferentially targeted by transposase (more accessible), we first align to chrM and filter these aligned reads out before aligning to human geome. \
We first align the PE1 and PE2 read to chrM, then we sort the file, use `samtools collate` to shuffle and group reads together with the same prefix, and use `samtools` and the `-f` option with the flag `0x4` to output the reads that do **NOT** align to the mitochondrial genome. \

```{r engine='bash', eval=TRUE, echo=T}
#! /bin/sh
#SBATCH --job-name=align_chrM.sh     
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 8                  
#SBATCH -p general           
#SBATCH --qos=general     
#SBATCH --mem=60G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o align_chrM.sh_%j.out
#SBATCH -e align_chrM.sh_%j.err

hostname
#export TMPDIR="${HOME}/temp" 

module load samtools/1.9
module load bowtie2

ncore=8
genome_index=/home/FCAM/ssun/Genome/humanchrM_bt2/chrM

#Align to chrM first and remove aligned and unpaired reads
for i in *_PE1_no_adapt.fastq
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1" '{print $1}')
    echo $name
    bowtie2 -p $ncore -x $genome_index -1 ${name}_PE1_no_adapt.fastq -2 ${name}_PE2_no_adapt.fastq | samtools sort -@ $ncore -n - | samtools collate -u - ${name} 2>&1 | tee ${name}_chrM_alignment.log
    samtools fastq -f 0x4 -1 ${name}_PE1.chrM.fastq -2 ${name}_PE2.chrM.fastq -0 /dev/null -s /dev/null -n ${name}.bam 
    rm ${name}.bam # remove intermediate file
done 
```


## Align to the human genome

The filterred reads (that remoed reads align to ChrM) are now aligning to the hg38.fa genome. \
`-1` and `-2` options for the respective paired-end fastq files. \
Then the output is piped to samtools to convert to bam, then sorted by name (`-n`) so paired end reads are adjacent in the file, then piped to `samtools fixmate` which adds information about the fragment length by comparing the PE1 and PE2 coordinates, then the files are sorted by coordinate, then piped to `samtools markdup` to remove duplicate reads. Duplicate reads have the same PE1 and PE2 ends. \
The fixmate step is necessary to pipe to markdup. \

```{r engine='bash', eval=TRUE, echo=T}
#! /bin/sh
#SBATCH --job-name=align_hg38.sh     
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 8                  
#SBATCH -p general           
#SBATCH --qos=general     
#SBATCH --mem=60G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o align_hg38.sh_%j.out
#SBATCH -e align_hg38.sh_%j.err

hostname
export TMPDIR="${HOME}/temp" 

module load samtools/1.16.1
module load bowtie2

ncore=8
genome_index=/home/FCAM/ssun/Genome/hg38_bt2/hg38

#Align to the hg38 genome and remove duplicates
for i in *_PE1_no_adapt.fastq
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1" '{print $1}')
    echo $name
    bowtie2 -p $ncore --maxins 800 -x $genome_index -1 ${name}_PE1.chrM.fastq -2 ${name}_PE2.chrM.fastq | samtools view -bS -f 0x2 - | samtools sort -@ $ncore -n - | samtools fixmate -m - - | samtools sort - | samtools markdup -s -r - ${name}.hg38.bam 2>&1 | tee ${name}_bowtie2_hg38.log #samtools view -bS -f 0x2 - keep only reads that are: properly paired and both reads are mapped
    #gzip ${name}_*.fastq
done
```

```{bash}
#250506 modification: get the unmapped sequences

#! /bin/sh
#SBATCH --job-name=align_hg38_unmapped.sh     
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 8                  
#SBATCH -p general           
#SBATCH --qos=general     
#SBATCH --mem=60G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o align_hg38_unmapped.sh_%j.out
#SBATCH -e align_hg38_unmapped.sh_%j.err

hostname
export TMPDIR="${HOME}/temp" 

module load samtools/1.16.1
module load bowtie2

ncore=8
genome_index=/home/FCAM/ssun/Genome/hg38_bt2/hg38

gunzip *DMSO*chrM.fastq.gz
for i in *DMSO*PE1.chrM.fastq
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1" '{print $1}')
    echo $name
bowtie2 -p $ncore --maxins 800 -x $genome_index \
  -1 ${name}_PE1.chrM.fastq -2 ${name}_PE2.chrM.fastq | \
samtools view -bS - | tee ${name}_all_reads.bam | \
samtools view -b -f 4 - > ${name}_unmapped.bam
 gzip ${name}_*.fastq
done
```


zip the intermediate files. \
```{r engine='bash', eval=TRUE, echo=T}
for i in *_PE1_no_adapt.fastq
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1" '{print $1}')
    echo $name
    gzip ${name}_*.fastq
done
```


## Convert aligned data to a bed file

`seqOutBias` mappability files generation: \
```{r engine='bash', eval=TRUE, echo=T}
#!/bin/bash
#SBATCH --job-name=seq_map.sh     
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 8                  
#SBATCH -p general         
#SBATCH --qos=general     
#SBATCH --mem=68G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o seq_map.sh_%j.out
#SBATCH -e seq_map.sh_%j.err

hostname

module load genometools/1.5.10
module load ucsc_genome/2012.05.22
module load rust
export PATH=$PATH:/home/FCAM/ssun/packages/
export TMPDIR="${HOME}/temp" 

genome=/home/FCAM/ssun/Genome/hg38.fa
seqOutBias seqtable ${genome} --read-size=40 2>&1 | tee seqOutBias_40.log
```

Convert BAM to bed for counting (previously created the tallymer with seqOutBias seqtable) \

```{r engine='bash', eval=TRUE, echo=T}
#!/bin/bash
#SBATCH --job-name=bamtobed.sh     
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 8                  
#SBATCH -p general         
#SBATCH --qos=general     
#SBATCH --mem=68G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o bamtobed.sh_%j.out
#SBATCH -e bamtobed.sh_%j.err

hostname

module load genometools/1.5.10
module load ucsc_genome/2012.05.22
module load rust
export PATH=$PATH:/home/FCAM/ssun/packages/
export TMPDIR="${HOME}/temp" 

read_size=40
tallymer=/home/FCAM/ssun/seqoutbias/hg38_rs40/hg38.tal_${read_size}.gtTxt.gz
table=/home/FCAM/ssun/seqoutbias/hg38_rs40/hg38_${read_size}.4.2.2.tbl

for i in *.hg38.bam
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F".hg38.bam" '{print $1}')
    echo $name
    seqOutBias scale $table ${name}.hg38.bam --tallymer=$tallymer --no-scale --custom-shift=4,-4 --read-size=${read_size} 2>&1 | tee ${name}_seqOutBias.log
    grep -v "random" ${name}.hg38_not_scaled.bed | grep -v "chrUn" | grep -v "chrEBV" | grep -v "alt" | sort -k1,1 -k2,2n > ${name}_tmp.txt && mv ${name}_tmp.txt ${name}_not_scaled.bed 
    rm ${name}.hg38_not_scaled.bed
done
```

## normalize to read depth

[normalize with scale factor or Deseq2 size factor] \

# Normalization & Genome Browser

Usually we combine replicates into a single track for visualization
and we compare the tracks between conditions. It is important that we
**read-depth normalize** (we are currently trying to change from read-depth 
normalize to deseq2 size-factor normalize) before we combine the signal from each
replicate. Otherwise, we would be weighting replicates differently;
for example, if a library is sequenced to twice the read depth and we
combine first then read depth normalize, then the more high coverage
data is weighted twice as much in the final visualization. All the
relevant track files are here: \

```{r}
#/home/FCAM/ssun/GATA3_ATAC_July2024/normalization

module load bedtools
module load samtools
module load genometools/1.5.10
module load ucsc_genome/2012.05.22

ncore=8
sizes=/home/FCAM/ssun/Genome/hg38.chrom.sizes


export TMPDIR="${HOME}/temp"
for i in ../cutadapt/*.hg38.bam
do
    name=$(echo $i | awk -F"dTAGGATA522_" '{print $NF}' | awk -F".hg38.bam" '{print $1}')
    prename=$(echo $i | awk -F"/" '{print $NF}' | awk -F".hg38.bam" '{print $1}')
    echo $name  
    #bowtie2 -p $ncore --maxins 800 -x $genome_index -1 ${name}_PE1_no_adapt.fastq -2 ${name}_PE2_no_adapt.fastq | samtools view -bS -f 0x2 - | samtools sort -@ $ncore -n | samtools fixmate -m - - | samtools sort - | samtools markdup -s -r - ${name}.hg38.bam 2>&1 | tee ${name}_bowtie2_hg38.log
    #seqOutFBias scale $table ${name}.hg38.bam --tallymer=$tallymer --shift-counts --no-scale --bw=${name}.bigWig --read-size=${readLength} 2>&1 | tee ${name}_seqOutBias.log
    samtools sort -@ $ncore -n $i -o ${name}.sorted.hg38.bam
    bedtools bamtobed -i ${name}.sorted.hg38.bam -bedpe > ${name}_bed12.bed
    awk '$1==$4 {print $0}' ${name}_bed12.bed | awk '{OFS="\t";} {print $1, $2, $6}' | awk '$1!="." && $3>$2 && (($3 - $2)<2000) {print $0}' | sort -k1,1 -k2,2n > ${name}_read_span.bed
    genomeCoverageBed -bg -i ${name}_read_span.bed -g $sizes > ${name}.bedGraph
    depth=`awk -F'\t' '{sum+=$5;}END{print sum;}' ../cutadapt/${prename}_not_scaled.bed`
    scaled=$(bc <<< "scale=3 ; 10000000 / $depth")
    echo $scaled
    awk -v scaled="$scaled" '{OFS="\t";} {print $1, $2, $3, $4*scaled}' ${name}.bedGraph > ${name}_normalized.bedGraph
    awk -v var="$name" 'BEGIN {  print "browser position chr11:5,289,521-5,291,937"; print "track type=bedGraph name=\"" var "\" description=\"" var "_bedGraph\" visibility=full autoScale=on alwaysZero=on color=0,125,0 windowingFunction=maximum"}  { print $0}' ${name}_normalized.bedGraph > ${name}_ucsc.bedGraph
    wigToBigWig -clip ${name}_normalized.bedGraph $sizes ${name}_normalized.bigWig
done

sizes=/home/FCAM/ssun/Genome/hg38.chrom.sizes

for i in ../cutadapt/*rep1.hg38.bam
do
    name=$(echo $i | awk -F"dTAGGATA522_" '{print $NF}' | awk -F"_rep1" '{print $1}')
    echo $name
    reps=$(ls ${name}_rep*normalized.bigWig | wc -w | bc)
    echo $reps
    files=$(ls ${name}_rep*normalized.bigWig)
    echo $files
    bigWigMerge ${name}_rep*normalized.bigWig tmp.bg
    scaleall=$(bc <<< "scale=4 ; 1.0 / $reps")
    echo scale:
    echo $scaleall
    awk -v scaleall="$scaleall" '{OFS="\t";} {print $1, $2, $3, $4*scaleall}' tmp.bg > ${name}_normalized.bedGraph
    rm tmp.bg
    wigToBigWig ${name}_normalized.bedGraph $sizes ${name}.bigWig
    awk -v var="$name" 'BEGIN {  print "browser position chr11:5,289,521-5,291,937"; print "track type=bedGraph name=\"" var "\" description=\"" var "_bedGraph\" visibility=full autoScale=on alwaysZero=on color=0,0,0"}  { print $0}' ${name}_normalized.bedGraph > ${name}_header_normalized.bedGraph
    gzip ${name}_header_normalized.bedGraph
done

wigToBigWig -clip ${name}_normalized.bedGraph $sizes ${name}_normalized.bigWig
```


## Single library (normalized) 

UCSC Genome Browser trackhub link for non-merged, normalized bigWig files: http://guertinlab.cam.uchc.edu/GATA3_ATAC_hub/hub.txt \


