---
title: "GATA3_ATAC_analysis"

author:
- Siyu Sun
- Michael J. Guertin
header-includes:
- \usepackage{color}
- \usepackage{float}
- \DeclareUnicodeCharacter{2212}{-}
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    toc: true
fontsize: 14pt
geometry: margin=1in
---

# ATAC-seq experiments

ATAC-seq measures genome-wide chromatin accessibility. 

# ATAC-seq experiments design

MCF7 cells with the dTAG-GATA3 (clone522) grow in full medium are treated under 2 conditions: 1) DMSO; 2) 30min dTAG-v1. \
MCF7 cells with the dTAG-GATA3 (clone522) grow in charcoal stripped medium (ER deprived) are treated under 2 conditions: 1) DMSO; 2) 30min dTAG-v1. \


# Raw Data Processing 

## Install required packages
Install `bigWig` and `latticeExtra` package \
```{r engine='R', eval=F, echo=T}
install.packages("devtools", quiet = TRUE)
library(devtools)
devtools::install_github('andrelmartins/bigWig',
              subdir='bigWig')
library(bigWig)

install.packages("DESeq2", quiet = TRUE)
install.packages("dplyr", quiet = TRUE)
```

Install `bedtools` \
```{r engine='bash', eval=F, echo=T}
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
brew install bedtools
```

Install `Biostrings` \
```{r engine='R', eval=F, echo=T}
if (!requireNamespace("Biostrings", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
  BiocManager::install("Biostrings")
}
```

Install `latticeExtra` \
```{r engine='R', eval=TRUE, echo=T}
install.packages("latticeExtra", quiet = TRUE)
```

## Renaming

```{r engine='bash', eval=TRUE, echo=T}
for i in *R1_001.fastq.gz
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_S" '{print $1}')
    cat ${name}*R1_001.fastq.gz > MCF7_dTAGGATAClone522_30min_${name}_PE1.fastq.gz
    cat ${name}*R2_001.fastq.gz > MCF7_dTAGGATAClone522_30min_${name}_PE2.fastq.gz    
done
```

## Cut off the adapter with `cutadapt`

`-j`: the number of cores; \
`-m` specifies the minimal length of a read to keep after adapter sequence removal; \
`-O` is the number of bases to trim off the end of the read if it overlaps with the adapter sequence. \
`-a` and `-A` options are the adapter sequences of the PE1 and PE2 reads. (Tn5 mosaic end sequence: CTGTCTCTTATACACATCT). \
`-o` (PE1) and `-p` (PE2) specfies the output file. \
The last two positional areguments are the input fastq files. We save the output to a log file. \

```{r engine='bash', eval=TRUE, echo=T}
#! /bin/sh
#SBATCH --job-name=GATA_atac_cutadapt.sh    
#SBATCH -N 1
#SBATCH -n 1
#SBATCH -c 8
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=32G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o cutadapt.sh_%j.out
#SBATCH -e cutadapt.sh_%j.err

hostname
module load cutadapt/3.5

for i in ../raw_fastq_renamed/*PE1.fastq
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1" '{print $1}')
    echo $name
    echo unzipping $i
    #gunzip $i
    echo unzipping ${name}_PE2.fastq.gz
    gunzip ../raw_fastq_renamed/${name}_PE2.fastq.gz
    cutadapt -a CTGTCTCTTATACACATCT -A CTGTCTCTTATACACATCT -j 8 -m 10 -O 1 -o ${name}_PE1_no_adapt.fastq -p ${name}_PE2_no_adapt.fastq ../raw_fastq_renamed/${name}_PE1.fastq ../raw_fastq_renamed/${name}_PE2.fastq 2>&1 | tee ${name}_cutadapt.log

done
```


# Alighment

## Filter reads that align to the mitochondrial chromosome
```{r engine='bash', eval=TRUE, echo=T}
#! /bin/sh
#SBATCH --job-name=humanchrM_bt2.sh     
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 4                 
#SBATCH -p general           
#SBATCH --qos=general     
#SBATCH --mem=10G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o humanchrM_bt2.sh_%j.out
#SBATCH -e humanchrM_bt2.sh_%j.err

module load genometools/1.5.10
module load bowtie2
export PATH=$PATH:/home/FCAM/ssun/packages

# build the mitochondrial genome
wget https://hgdownload.cse.ucsc.edu/goldenpath/hg38/chromosomes/chrM.fa.gz
gunzip chrM.fa.gz
bowtie2-build chrM.fa chrM
```

Since mitochondrial DNA is preferentially targeted by transposase (more accessible), we first align to chrM and filter these aligned reads out before aligning to human genome. \
We first align the PE1 and PE2 read to chrM, then we sort the file, use `samtools collate` to shuffle and group reads together with the same prefix, and use `samtools` and the `-f` option with the flag `0x4` to output the reads that do **NOT** align to the mitochondrial genome. \

```{r engine='bash', eval=TRUE, echo=T}
#! /bin/sh
#SBATCH --job-name=align_chrM.sh     
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 8                  
#SBATCH -p general           
#SBATCH --qos=general     
#SBATCH --mem=60G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o align_chrM.sh_%j.out
#SBATCH -e align_chrM.sh_%j.err

hostname
#export TMPDIR="${HOME}/temp" 

module load samtools/1.9
module load bowtie2

ncore=8
genome_index=/home/FCAM/ssun/Genome/humanchrM_bt2/chrM

#Align to chrM first and remove aligned and unpaired reads
for i in *_PE1_no_adapt.fastq
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1" '{print $1}')
    echo $name
    bowtie2 -p $ncore -x $genome_index -1 ${name}_PE1_no_adapt.fastq -2 ${name}_PE2_no_adapt.fastq | samtools sort -@ $ncore -n - | samtools collate -u - ${name} 2>&1 | tee ${name}_chrM_alignment.log
    samtools fastq -f 0x4 -1 ${name}_PE1.chrM.fastq -2 ${name}_PE2.chrM.fastq -0 /dev/null -s /dev/null -n ${name}.bam 
    rm ${name}.bam # remove intermediate file
done 
```


## Align to the human genome

The filterred reads (that have removed reads align to ChrM) are now aligning to the hg38.fa genome. \
`-1` and `-2` options for the respective paired-end fastq files. \
Then the output is piped to samtools to convert to bam, then sorted by name (`-n`) so paired end reads are adjacent in the file, then piped to `samtools fixmate` which adds information about the fragment length by comparing the PE1 and PE2 coordinates, then the files are sorted by coordinate, then piped to `samtools markdup` to remove duplicate reads. Duplicate reads have the same PE1 and PE2 ends. \
The fixmate step is necessary to pipe to markdup. \

```{r engine='bash', eval=TRUE, echo=T}
#! /bin/sh
#SBATCH --job-name=align_hg38.sh     
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 8                  
#SBATCH -p general           
#SBATCH --qos=general     
#SBATCH --mem=60G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o align_hg38.sh_%j.out
#SBATCH -e align_hg38.sh_%j.err

hostname
export TMPDIR="${HOME}/temp" 

module load samtools/1.16.1
module load bowtie2

ncore=8
genome_index=/home/FCAM/ssun/Genome/hg38_bt2/hg38

#Align to the hg38 genome and remove duplicates
for i in *_PE1_no_adapt.fastq
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1" '{print $1}')
    echo $name
    bowtie2 -p $ncore --maxins 800 -x $genome_index -1 ${name}_PE1.chrM.fastq -2 ${name}_PE2.chrM.fastq | samtools view -bS -f 0x2 - | samtools sort -@ $ncore -n - | samtools fixmate -m - - | samtools sort - | samtools markdup -s -r - ${name}.hg38.bam 2>&1 | tee ${name}_bowtie2_hg38.log #samtools view -bS -f 0x2 - keep only reads that are: properly paired and both reads are mapped
    #gzip ${name}_*.fastq
done
```

```{bash}
#250506 modification: get the unmapped sequences

#! /bin/sh
#SBATCH --job-name=align_hg38_unmapped.sh     
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 8                  
#SBATCH -p general           
#SBATCH --qos=general     
#SBATCH --mem=60G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o align_hg38_unmapped.sh_%j.out
#SBATCH -e align_hg38_unmapped.sh_%j.err

hostname
export TMPDIR="${HOME}/temp" 

module load samtools/1.16.1
module load bowtie2

ncore=8
genome_index=/home/FCAM/ssun/Genome/hg38_bt2/hg38

gunzip *DMSO*chrM.fastq.gz
for i in *DMSO*PE1.chrM.fastq
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1" '{print $1}')
    echo $name
bowtie2 -p $ncore --maxins 800 -x $genome_index \
  -1 ${name}_PE1.chrM.fastq -2 ${name}_PE2.chrM.fastq | \
samtools view -bS - | tee ${name}_all_reads.bam | \
samtools view -b -f 4 - > ${name}_unmapped.bam
 gzip ${name}_*.fastq
done
```


zip the intermediate files. \
```{r engine='bash', eval=TRUE, echo=T}
for i in *_PE1_no_adapt.fastq
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1" '{print $1}')
    echo $name
    gzip ${name}_*.fastq
done
```


## Convert aligned data to a bed file

We used `seqOutBias` to convert the data into BED format. The `--no-scale` flag was included to prevent correction for enzymatic sequence bias in chromatin accessibility assays, and `--custom-shift=4,-4` was used to specify the center of the transposase interaction site. The `grep` step removes incomplete contigs and the Epstein-Barr virus genome. \

`seqOutBias` mappability files generation: \
```{r engine='bash', eval=TRUE, echo=T}
#!/bin/bash
#SBATCH --job-name=seq_map.sh     
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 8                  
#SBATCH -p general         
#SBATCH --qos=general     
#SBATCH --mem=68G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o seq_map.sh_%j.out
#SBATCH -e seq_map.sh_%j.err

hostname

module load genometools/1.5.10
module load ucsc_genome/2012.05.22
module load rust
export PATH=$PATH:/home/FCAM/ssun/packages/
export TMPDIR="${HOME}/temp" 

genome=/home/FCAM/ssun/Genome/hg38.fa
seqOutBias seqtable ${genome} --read-size=40 2>&1 | tee seqOutBias_40.log
```

Convert BAM to bed for counting (previously created the tallymer with seqOutBias seqtable) \

```{r engine='bash', eval=TRUE, echo=T}
#!/bin/bash
#SBATCH --job-name=bamtobed.sh     
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 8                  
#SBATCH -p general         
#SBATCH --qos=general     
#SBATCH --mem=68G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o bamtobed.sh_%j.out
#SBATCH -e bamtobed.sh_%j.err

hostname

module load genometools/1.5.10
module load ucsc_genome/2012.05.22
module load rust
export PATH=$PATH:/home/FCAM/ssun/packages/
export TMPDIR="${HOME}/temp" 

read_size=40
tallymer=/home/FCAM/ssun/seqoutbias/hg38_rs40/hg38.tal_${read_size}.gtTxt.gz
table=/home/FCAM/ssun/seqoutbias/hg38_rs40/hg38_${read_size}.4.2.2.tbl

for i in *.hg38.bam
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F".hg38.bam" '{print $1}')
    echo $name
    seqOutBias scale $table ${name}.hg38.bam --tallymer=$tallymer --no-scale --custom-shift=4,-4 --read-size=${read_size} 2>&1 | tee ${name}_seqOutBias.log
    grep -v "random" ${name}.hg38_not_scaled.bed | grep -v "chrUn" | grep -v "chrEBV" | grep -v "alt" | sort -k1,1 -k2,2n > ${name}_tmp.txt && mv ${name}_tmp.txt ${name}_not_scaled.bed 
    rm ${name}.hg38_not_scaled.bed
done
```


# Normalization & Genome Browser

Usually we combine replicates into a single track for visualization
and we compare the tracks between conditions. It is important that we
**read-depth normalize** (we are currently trying to change from read-depth 
normalize to deseq2 size-factor normalize) before we combine the signal from each
replicate. Otherwise, we would be weighting replicates differently;
for example, if a library is sequenced to twice the read depth and we
combine first then read depth normalize, then the more high coverage
data is weighted twice as much in the final visualization. All the
relevant track files are here: \

## normalize to read depth

[normalize with scale factor or Deseq2 size factor] \


```{r}
#/home/FCAM/ssun/GATA3_ATAC_July2024/normalization

module load bedtools
module load samtools
module load genometools/1.5.10
module load ucsc_genome/2012.05.22

ncore=8
sizes=/home/FCAM/ssun/Genome/hg38.chrom.sizes


export TMPDIR="${HOME}/temp"
for i in ../cutadapt/*.hg38.bam
do
    name=$(echo $i | awk -F"dTAGGATA522_" '{print $NF}' | awk -F".hg38.bam" '{print $1}')
    prename=$(echo $i | awk -F"/" '{print $NF}' | awk -F".hg38.bam" '{print $1}')
    echo $name  
    #bowtie2 -p $ncore --maxins 800 -x $genome_index -1 ${name}_PE1_no_adapt.fastq -2 ${name}_PE2_no_adapt.fastq | samtools view -bS -f 0x2 - | samtools sort -@ $ncore -n | samtools fixmate -m - - | samtools sort - | samtools markdup -s -r - ${name}.hg38.bam 2>&1 | tee ${name}_bowtie2_hg38.log
    #seqOutFBias scale $table ${name}.hg38.bam --tallymer=$tallymer --shift-counts --no-scale --bw=${name}.bigWig --read-size=${readLength} 2>&1 | tee ${name}_seqOutBias.log
    samtools sort -@ $ncore -n $i -o ${name}.sorted.hg38.bam
    bedtools bamtobed -i ${name}.sorted.hg38.bam -bedpe > ${name}_bed12.bed
    awk '$1==$4 {print $0}' ${name}_bed12.bed | awk '{OFS="\t";} {print $1, $2, $6}' | awk '$1!="." && $3>$2 && (($3 - $2)<2000) {print $0}' | sort -k1,1 -k2,2n > ${name}_read_span.bed
    genomeCoverageBed -bg -i ${name}_read_span.bed -g $sizes > ${name}.bedGraph
    depth=`awk -F'\t' '{sum+=$5;}END{print sum;}' ../cutadapt/${prename}_not_scaled.bed`
    scaled=$(bc <<< "scale=3 ; 10000000 / $depth")
    echo $scaled
    awk -v scaled="$scaled" '{OFS="\t";} {print $1, $2, $3, $4*scaled}' ${name}.bedGraph > ${name}_normalized.bedGraph
    awk -v var="$name" 'BEGIN {  print "browser position chr11:5,289,521-5,291,937"; print "track type=bedGraph name=\"" var "\" description=\"" var "_bedGraph\" visibility=full autoScale=on alwaysZero=on color=0,125,0 windowingFunction=maximum"}  { print $0}' ${name}_normalized.bedGraph > ${name}_ucsc.bedGraph
    wigToBigWig -clip ${name}_normalized.bedGraph $sizes ${name}_normalized.bigWig
done

sizes=/home/FCAM/ssun/Genome/hg38.chrom.sizes

for i in ../cutadapt/*rep1.hg38.bam
do
    name=$(echo $i | awk -F"dTAGGATA522_" '{print $NF}' | awk -F"_rep1" '{print $1}')
    echo $name
    reps=$(ls ${name}_rep*normalized.bigWig | wc -w | bc)
    echo $reps
    files=$(ls ${name}_rep*normalized.bigWig)
    echo $files
    bigWigMerge ${name}_rep*normalized.bigWig tmp.bg
    scaleall=$(bc <<< "scale=4 ; 1.0 / $reps")
    echo scale:
    echo $scaleall
    awk -v scaleall="$scaleall" '{OFS="\t";} {print $1, $2, $3, $4*scaleall}' tmp.bg > ${name}_normalized.bedGraph
    rm tmp.bg
    wigToBigWig ${name}_normalized.bedGraph $sizes ${name}.bigWig
    awk -v var="$name" 'BEGIN {  print "browser position chr11:5,289,521-5,291,937"; print "track type=bedGraph name=\"" var "\" description=\"" var "_bedGraph\" visibility=full autoScale=on alwaysZero=on color=0,0,0"}  { print $0}' ${name}_normalized.bedGraph > ${name}_header_normalized.bedGraph
    gzip ${name}_header_normalized.bedGraph
done

wigToBigWig -clip ${name}_normalized.bedGraph $sizes ${name}_normalized.bigWig
```


## Single library (normalized) 

UCSC Genome Browser trackhub link for non-merged, normalized bigWig files: http://guertinlab.cam.uchc.edu/GATA3_ATAC_hub/hub.txt \


# Call peaks

## MACS2

We use `macs2` to call peaks using all the `*.hg38.bam` files. \
We already removed replicate duplicates, so any true duplicates arose independently. We could call peaks in individual replicates and take the union or intersect or there are many other ways to call peaks. Since we don’t have an input or control file (we don't have input DNA or DNA pulled down with an unspecific antibody as in ChIP-seq; The only proper control would probably be to use the Tn on "naked" genomic DNA. But then you need quiet many reads to get a proper signal due to the size of mammalian genomes, so the cost/effect ratio is simply not economic so nobody routinely does it...) and most biologically relevant peaks will be found by most peak callers,we should find the important peaks where we have the power to detect changes in accessibility. \

*the early ATAC papers (maybe the original one and the one on the NucleoATAC software). They show that the intrinsic cutting preference/bias of the transposon is minimal if I remember correctly. \
 
`--nomodel` \
While on, MACS will bypass building the shifting model. \
`--extsize` \
While '--nomodel' is set, MACS uses this parameter to extend reads in 5'->3' direction to fix-sized fragments. For example, if the size of binding region for your transcription factor is 200 bp, and you want to bypass the model building by MACS, this parameter can be set as 200. This option is only valid when --nomodel is set or when MACS fails to build model and --fix-bimodal is on. \
`--shift` \
Note, this is NOT the legacy --shiftsize option which is replaced by --extsize! You can set an arbitrary shift in bp here. Please Use discretion while setting it other than default value (0). When --nomodel is set, MACS will use this value to move cutting ends (5') then apply --extsize from 5' to 3' direction to extend them to fragments. When this value is negative, ends will be moved toward 3'->5' direction, otherwise 5'->3' direction. Recommended to keep it as default 0 for ChIP-Seq datasets, or -1 * half of EXTSIZE together with --extsize option for detecting enriched cutting loci such as certain DNAseI-Seq datasets. Note, you can't set values other than 0 if format is BAMPE or BEDPE for paired-end data. Default is 0. `

```{r engine='bash', eval=F, echo=TRUE}
#! /bin/sh
#SBATCH --job-name=gata3_atac_peak_calling.sh    
#SBATCH -N 1
#SBATCH -n 1
#SBATCH -c 8
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=32G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o gata3_atac_pesk_calling.sh_%j.out
#SBATCH -e gata3_atac_peak_calling.sh_%j.err

hostname
module load macs2

mkdir temp_macs
directory=/home/FCAM/ssun/GATA3_ATAC_July2024/cutadapt/
# call peaks with MACS2
macs2 callpeak --call-summits -t ${directory}*.hg38.bam -n GATA3_degron_ATAC -g hs -q 0.01 --keep-dup all -f BAM --nomodel --shift -100 --extsize 200 --tempdir temp_macs

#wc -l GATA3_degron_ATAC_summits.bed
#259635 
```



## Removing peaks on contigs and within blacklisted regions

```{r engine='bash', eval=F, echo=TRUE}
wget https://github.com/Boyle-Lab/Blacklist/raw/master/lists/hg38-blacklist.v2.bed.gz
gunzip hg38-blacklist.v2.bed.gz
blacklist=hg38-blacklist.v2.bed
sizes=/home/FCAM/ssun/Genome_pro/hg38.chrom.sizes
name=GATA3_degron_ATAC

module load bedtools
grep -v "random" ${name}_summits.bed | grep -v "chrUn" | grep -v "chrEBV" | grep -v "chrM" | grep -v "alt" | intersectBed -v -a stdin -b $blacklist > ${name}_tmp.txt #247838
mv ${name}_tmp.txt ${name}_summits.bed

slopBed -b 200 -i ${name}_summits.bed -g $sizes > ${name}_summit_400window.bed #247838
```


# Differential reads counts with mapBed & FRip 

`mapBed` from bedtools counts reads in regions with two bed file inputs. We will counts reads in the 400 base window summit file and merge these into a file with experiment name columns and genomic interval named rows. \

FRiP (Fraction of Reads in Peaks), is a common quality measurement for ChIP-seq experiment. It calculates the percentage of all reads that fall in peak regions mapped to the genome. It is important that we always use a universal peak set to calculate the FRiP score for each replicates, then compare between replicates. Usually, we will use the peak set that we generated previously from MACS3, since in that step we have called peaks using all replicates in bam format for the same ChIP factor. \

```{r, engine='bash', eval=F, echo=TRUE}
module load bedtools
module load samtools/1.12
peaks=/home/FCAM/ssun/GATA3_ATAC_July2024/peak_call/GATA3_degron_ATAC_summit_400window.bed

sort -k1,1 -k2,2n $peaks > GATA3_degron_ATAC_summit_400window_sorted.bed

for i in /home/FCAM/ssun/GATA3_ATAC_July2024/cutadapt/*_not_scaled.bed
do
    nm=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_not_scaled.bed" '{print $1}')
    echo $nm
    sort -k1,1 -k2,2n $i > ${nm}_sorted.bed
    mapBed -null '0' -a GATA3_degron_ATAC_summit_400window_sorted.bed -b ${nm}_sorted.bed > ${nm}_peak_counts.txt #count 400bp window sum reads
done

#head MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_CS_DMSO_rep1_peak_counts.txt
#chr1	804681	805082	GATA3_degron_ATAC_peak_78	892.55103	27
#chr1	820107	820508	GATA3_degron_ATAC_peak_79	11.52228	7
#chr1	826823	827224	GATA3_degron_ATAC_peak_80a	99.70275	41
#chr1	827482	827883	GATA3_degron_ATAC_peak_80b	537.96899	89
#chr1	832118	832519	GATA3_degron_ATAC_peak_81	12.95156	11

for i in *_peak_counts.txt
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_peak_counts.txt" '{print $1}')
    echo $name > ${name}_FRiP.txt
    awk '{print $NF}' ${name}_peak_counts.txt > ${name}_peak_counts_only.txt
    RiP=$(awk '{sum+=$1;} END{print sum;}' ${name}_peak_counts_only.txt) # count the reads in each peak region (from the universal set)
    echo $name | cat - ${name}_peak_counts_only.txt > ${name}_peak_counts.txt
    aligned_reads=$(samtools view -c -f 0x2 /home/FCAM/ssun/GATA3_ATAC_July2024/cutadapt/${name}.hg38.bam) # count the total aligned reads (-f 0x2: read mapped in proper pair)
    FRiP=$(echo "scale=2 ; $RiP / $aligned_reads" | bc)
    echo $FRiP >> ${name}_FRiP.txt
    rm ${name}_peak_counts_only.txt
done

echo -e "chr\tstart\tend\tname\tqvalue" | cat - GATA3_degron_ATAC_summit_400window_sorted.bed | paste -d'\t' - *peak_counts.txt > Combined_GATA_peak_counts.txt
echo -e "Name" > rowname.txt
echo -e "FrIP" >> rowname.txt
paste -d'\t' rowname.txt *FRiP.txt > Combined_GATA_FRiP.txt
rm rowname.txt
```

```{r, engine='R', eval=F, echo=TRUE}
library(lattice)
frip.df=read.csv("GATA_ATAC_peak_FRiP.csv")
frip.df$supp=sapply(strsplit(frip.df$library, '_r'), '[', 1)

frip.df$supp=factor(frip.df$supp, levels=c("CS_DMSO", "CS_dTAG", "Full_DMSO", "Full_dTAG"))

pdf('barplot_GATA3_ATAC_peak_FRiP.pdf',width=8,height=8)
print(barchart(library~FRiP,
         group=supp,
         stack=T,
         data = frip.df,
         xlim=c(0,1),
         #auto.key=list(space="right"),
         #scales = list(x = list(rot = 45)),
         ylab = "ATAC-seq libraries",
         xlab = "FRiP",
         main ="MCF7 GATA3 ATAC peak FRiP",
         par.settings = list(superpose.polygon=list(col=c("pink","lightgreen", "lightblue","violet"), border="grey")),
         col=c("pink","lightgreen", "lightblue","violet")
))
dev.off()
```





# Differential accessibility

```{r, engine='bash', eval=F, echo=TRUE}
#cd /home/FCAM/ssun/GATA3_ATAC_July2024/ATAC_analysis/240718_deseq2

cp ../../FRIP/Combined_GATA_peak_counts.txt .
# head Combined_GATA_peak_counts.txt

#chr	start	end	name	qvalue	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_CS_DMSO_rep1	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_CS_DMSO_rep2	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_CS_DMSO_rep3	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_CS_DMSO_rep4	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_CS_dTAG_rep1	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_CS_dTAG_rep2	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_CS_dTAG_rep3	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_CS_dTAG_rep4	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_Full_DMSO_rep1	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_Full_DMSO_rep2	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_Full_DMSO_rep3	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_Full_DMSO_rep4	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_Full_dTAG_rep1	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_Full_dTAG_rep2	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_Full_dTAG_rep3	MCF7_dTAGGATAClone522_30min_MCF7_dTAGGATA522_Full_dTAG_rep4
#chr1	804681	805082	GATA3_degron_ATAC_peak_78	892.55103	27	56	31	39	93	29	43	18	80	35	66	67	65	77	29	74
#chr1	820107	820508	GATA3_degron_ATAC_peak_79	11.52228	7	31	4	10	7	17	13	16	19	17	18	22
```

load required libraries and functions: \
```{r, engine='R', eval=F, echo=TRUE}
# libraries
library(DESeq2)
library(lattice)
library(dplyr)
library(ggplot2)
library(limma)
library(MatchIt)
```

## full data sets (16 libraries)

```{r, engine='R', eval=F, echo=TRUE}
# data processing and plotting
x = read.table('Combined_GATA_peak_counts.txt', sep = "\t", check.names=FALSE, header=TRUE)
rownames(x) = paste0(x[,1], ":", x[,2], "-", x[,3])
peak_name = x[,4]
peak_qvalue = x[,5]
x = x[,-c(1:5)]

colnames(x) = sapply(strsplit(colnames(x), "dTAGGATA522_"), "[", 2)

#head(x)
#                   CS_DMSO_rep1 CS_DMSO_rep2 CS_DMSO_rep3 CS_DMSO_rep4 ...
#chr1:804681-805082           27           56           31           39 ...
#chr1:820107-820508            7            6            7            8 ...
#chr1:826823-827224           41          109           64           93 ...
#chr1:827482-827883           89          194          129          146 ...
#chr1:832118-832519           11           16            6           16 ...
#chr1:845620-846021          152          350          238          272 ...
```


## DESeq2 on full-medium samples (8)

```{r, engine='R', eval=F, echo=TRUE}
x_full=x[, 9:16]
sample.conditions = factor(sapply(strsplit(colnames(x_full), '_r'), '[', 1), levels=c("Full_DMSO", "Full_dTAG"))
#rep = factor(sapply(strsplit(colnames(x_full), 'rep'), '[', 2)) 
#deseq.counts.table = DESeqDataSetFromMatrix(countData = x_full,
#                colData = cbind.data.frame(sample.conditions, rep), 
#                design = ~ rep + sample.conditions)
deseq.counts.table = DESeqDataSetFromMatrix(countData = x_full,
                                            colData = DataFrame(sample.conditions), 
                                            design = ~ sample.conditions)
deseq.counts.table 

dds = DESeq(deseq.counts.table)
dds
resultsNames(dds)
#[1] "Intercept"                               
#[2] "sample.conditions_Full_dTAG_vs_Full_DMSO"
save(dds, file = "GATA3_ATAC_Full_dds.Rdata")

normalized.counts = counts(dds, normalized=TRUE)
rld = rlog(dds, blind=TRUE)
save(rld, file = "GATA3_ATAC_Full_rld.Rdata")
```

### plot principle components 
```{r, engine='R', eval=F, echo=TRUE}
pca.plot = plotPCA(rld, intgroup="sample.conditions", returnData=TRUE)
pca.plot 
plotPCAlattice(pca.plot, file = 'ATAC_PCA_plot_Full_30min_GATA3_degradation.pdf')
```

### MA plot


```{r engine='R', eval=F, echo=TRUE}
#load("GATA3_ATAC_Full_dds.Rdata")
res_test = results(dds)
sum(res_test$padj < 0.1 & !is.na(res_test$padj))
#35002
sum(res_test$padj < 0.01 & !is.na(res_test$padj))
#18475
```

When we have many genes/regions to match, matchit generates error: `Error in `matchit()`: ! (from optmatch) result would exceed 2^31-1 bytes`. Thus, we need to modify the function to get around of it. \

**function2** (solve the matchit capacity): \
```{r, engine='R', eval=F, echo=TRUE}
# here we add a div parameter to the function to help subset large data to accelerate matchit performance
categorize.DEseq.df.mod <- function(DE.results, unchanged.padj = 0.25, changed.padj = 0.1, div=15, match.method="optimal") {
    DE.results.lattice = DE.results
    DE.results.lattice$category = "All Other Genes"
    
    #find genes/regions match to activated
    activated = DE.results.lattice[DE.results.lattice$padj < changed.padj & !is.na(DE.results.lattice$padj) & DE.results.lattice$log2FoldChange > 0,]
    #subset the data
    df=activated
    rows_per_subset=ceiling(nrow(df) / div)
    split_df <- lapply(1:div, function(i) {start_row <- (i - 1) * rows_per_subset + 1
    end_row <- min(i * rows_per_subset, nrow(df))
    df[start_row:end_row, ]
})
    for (i in 1:div) {
      activated_subset=split_df[[i]]
      unchanged.act = DE.results.lattice[!is.na(DE.results.lattice$padj) & DE.results.lattice$padj > 0.1 & abs(DE.results.lattice$log2FoldChange) < unchanged.padj & DE.results.lattice$category == "All Other Genes",]
      unchanged.act$treatment = 0
      activated_subset$treatment = 1
      set.seed(123) # for reproducibility
      unchanged.act_subset=unchanged.act[sample(nrow(unchanged.act), 1000), ]  # randomly select 1000 unchanged.act data to match 
      df.deseq.effects.lattice = rbind(unchanged.act_subset, activated_subset)
      out = matchit(treatment ~ baseMean, data = df.deseq.effects.lattice, method = match.method, ratio = 1)
      unchanged.act = df.deseq.effects.lattice[rownames(df.deseq.effects.lattice) %in% out$match.matrix,]
      matching_rows <- intersect(rownames(DE.results.lattice), rownames(unchanged.act))
      for (row_name in matching_rows) {
        DE.results.lattice[row_name, "category"] <- "Matched to Up"
    }
    }
    
    #find genes/regions match to repressed
    repressed = DE.results.lattice[DE.results.lattice$padj < changed.padj & !is.na(DE.results.lattice$padj) & DE.results.lattice$log2FoldChange < 0,]
    #subset the data
    df=repressed
    rows_per_subset=ceiling(nrow(df) / div)
    split_df <- lapply(1:div, function(i) {start_row <- (i - 1) * rows_per_subset + 1
    end_row <- min(i * rows_per_subset, nrow(df))
    df[start_row:end_row, ]
})
    for (i in 1:div) {
      repressed_subset=split_df[[i]]
      unchanged.rep = DE.results.lattice[!is.na(DE.results.lattice$padj) & DE.results.lattice$padj > 0.1 & abs(DE.results.lattice$log2FoldChange) < unchanged.padj & DE.results.lattice$category == "All Other Genes",]
      unchanged.rep$treatment = 0
      repressed_subset$treatment = 1
      set.seed(123) # for reproducibility
      unchanged.rep_subset=unchanged.rep[sample(nrow(unchanged.rep), 1000), ]  # randomly select 1000 unchanged.act data to match 
      df.deseq.effects.lattice = rbind(unchanged.rep_subset, repressed_subset)
      out = matchit(treatment ~ baseMean, data = df.deseq.effects.lattice, method = match.method, ratio = 1)
      unchanged.rep = df.deseq.effects.lattice[rownames(df.deseq.effects.lattice) %in% out$match.matrix,]
      matching_rows <- intersect(rownames(DE.results.lattice), rownames(unchanged.rep))
      for (row_name in matching_rows) {
        DE.results.lattice[row_name, "category"] <- "Matched to Down"
    }
    }
    
      # determine genes/regions that are activated or repressed
      DE.results.lattice$category <- ifelse(
        DE.results.lattice$padj < changed.padj & !is.na(DE.results.lattice$padj) & DE.results.lattice$log2FoldChange > 0,
        "Up",
        DE.results.lattice$category
        )
      DE.results.lattice$category <- ifelse(
        DE.results.lattice$padj < changed.padj & !is.na(DE.results.lattice$padj) & DE.results.lattice$log2FoldChange < 0,
        "Down",
        DE.results.lattice$category
        )
    

    return(DE.results.lattice)
}
```

```{r, engine='R', eval=F, echo=TRUE}
#load("GATA3_ATAC_Full_dds.Rdata")
DE.results = results(dds)
head(DE.results)

DE.results.lattice = categorize.DEseq.df.mod(DE.results, unchanged.padj = 0.03, changed.padj = 0.01, div=50, match.method= "optimal")
nrow(DE.results.lattice)
#[1] 247838
nrow(DE.results.lattice[DE.results.lattice$category=="Up",])
#[1] 4418
nrow(DE.results.lattice[DE.results.lattice$category=="Down",])
#[1] 14057


save.image('240718_GATA3_ATAC_Full_DE_results.Rdata')
```


draw MA plot: \

```{r, engine='R', eval=F, echo=TRUE}
ma.plot.lattice <- function(ma.df, filename = 'file.name', 
                            title.main = "Differential ATAC-seq Accessibility",
                            col = c("grey80", "grey20" , "grey20", "#2290cf", "#ce228e")){
    desired_order <- c("All Other Genes", "Matched to UP", "Matched to Down", "Down", "Up")
    ma.df$category <- factor(ma.df$category, levels = desired_order)
  pdf(paste("MA_plot_", filename, ".pdf", sep=''), 
      useDingbats = FALSE, width=3.83, height=3.83);
  print(xyplot(ma.df$log2FoldChange ~ log(ma.df$baseMean, base=10),
               groups=ma.df$category,
               col= col,
                main=title.main, scales="free", aspect=1, pch=20, cex=0.5,
               ylab=expression("log"[2]~"ATAC-seq change"), 
               xlab=expression("log"[10]~"Mean of Normalized Counts"),
               par.settings=list(par.xlab.text=list(cex=1.1,font=2), 
                                 par.ylab.text=list(cex=1.1,font=2))));
  dev.off()
}

ma.plot.lattice(DE.results.lattice, filename = "ATAC_GATA3_MCF7_in_Full_medium_30min_dTAG_accessiblity_MA")
```

**get example ATAC-seq decreased region** \

```{r}
#/home/FCAM/ssun/GATA3_ATAC_July2024/ATAC_analysis/240718_deseq2
load("240718_GATA3_ATAC_Full_DE_results.Rdata")

head(DE.results.lattice)

DE.results.lattice[DE.results.lattice$log2FoldChange < -1.7 & log10(DE.results.lattice$baseMean) > 3.4 & DE.results.lattice$category=="Down", ]
#chr17:61219408-61219809
```


### Save differential peaks and matched peaks to .bed file

```{r, engine='R', eval=F, echo=TRUE}
options(scipen = 99) #avoid scientific notation when writing output
matched.all = DE.results.lattice[DE.results.lattice$category == 
                                 'Matched to Down' |
                                DE.results.lattice$category == 
                                 'Matched to Up',]

repressed.all = DE.results.lattice[DE.results.lattice$category == 
                                             'Down',]

activated.all = DE.results.lattice[DE.results.lattice$category == 
                                             'Up',]

chr = sapply(strsplit(rownames(activated.all), ":"), "[", 1)
rnge = sapply(strsplit(rownames(activated.all), ":"), "[", 2)
start = as.numeric(sapply(strsplit(rnge, "-"), "[", 1)) +100
end = as.numeric(sapply(strsplit(rnge, "-"), "[", 2)) -100 

write.table(cbind(chr, start, end), file = "Full_GATA3_degron_ATAC_up_200bp.bed", quote = FALSE,
col.names =FALSE, row.names=FALSE, sep = "\t")


chr = sapply(strsplit(rownames(repressed.all), ":"), "[", 1)
rnge = sapply(strsplit(rownames(repressed.all), ":"), "[", 2)
start = as.numeric(sapply(strsplit(rnge, "-"), "[", 1)) +100
end = as.numeric(sapply(strsplit(rnge, "-"), "[", 2)) -100

write.table(cbind(chr, start, end), file = "Full_GATA3_degron_ATAC_down_200bp.bed", quote = FALSE,
col.names =FALSE, row.names=FALSE, sep = "\t")

chr = sapply(strsplit(rownames(matched.all), ":"), "[", 1)
rnge = sapply(strsplit(rownames(matched.all), ":"), "[", 2)
start = as.numeric(sapply(strsplit(rnge, "-"), "[", 1)) +100
end = as.numeric(sapply(strsplit(rnge, "-"), "[", 2)) -100


matched.file = cbind(chr, start, end)
matched.file = unique(matched.file)

write.table(matched.file, file = "Full_GATA3_degron_ATAC_matched_200bp.bed", quote = FALSE,
col.names =FALSE, row.names=FALSE, sep = "\t")
```


# Overlapped ChIP-seq and ATAC-seq peaks

We can use `intersectBed` to find ATAC-seq peaks (up/down/match) that overlap with ChIP-seq peaks. \
## GATA3 ChIP-peaks
```{r, engine='bash', eval=F, echo=TRUE}
#cd /home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/peak_call/
module load bedtools
sizes=/home/FCAM/ssun/Genome/hg38.chrom.sizes
slopBed -b 100 -i GATA_ChIP_summits_final.bed -g $sizes  | sort -k1,1 -k2,2n > GATA_ChIP_summit_200window.bed
```

## GATA3 ATAC-peaks
did not use these:
CS_GATA3_degron_ATAC_down_200bp.bed
CS_GATA3_degron_ATAC_matched_200bp.bed
CS_GATA3_degron_ATAC_up_200bp.bed

used these:
Full_GATA3_degron_ATAC_down_200bp.bed
Full_GATA3_degron_ATAC_matched_200bp.bed
Full_GATA3_degron_ATAC_up_200bp.bed

```{r, engine='bash', eval=F, echo=TRUE}
module load bedtools
ChIP_file=/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/peak_call/GATA_ChIP_summit_200window.bed

touch overlap_peak.txt
echo -e "medium\tsupp\toverlap_peaks\tfraction" > overlap_peak.txt


for i in /home/FCAM/ssun/GATA3_ATAC_July2024/ATAC_analysis/240718_deseq2/Full*200bp.bed
do
  nm=$(echo $i | awk -F"/" '{print $NF}' | awk -F".bed" '{print $1}')
  echo $nm
  sort -k1,1 -k2,2n $i > ${nm}_sorted.bed
  name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"Full_GATA3_degron_ATAC_" '{print $2}' | awk -F"_200bp.bed" '{print $1}')
  atac_peak=$(wc -l $i | awk '{print $1}')
  nonoverlap_peak=$(intersectBed -v -a $i -b $ChIP_file | wc -l | awk '{print $1}')
  overlap_peak=$(echo "scale=2 ; $atac_peak - $nonoverlap_peak" | bc)
  fraction=$(echo "scale=2 ; $overlap_peak / $atac_peak" | bc)
  echo -e "Full\t$name\t$overlap_peak\t$fraction" >> overlap_peak.txt
done

#cd overlap2
cat overlap_peak.txt 
#medium	supp	overlap_peaks	fraction
#CS	down	6741	.95
#CS	matched_Down	2044	.29
#CS	matched_Up	653	.41
#CS	up	677	.42
#Full	down	13025	.92
#Full	matched_Down	3171	.22
#Full	matched_Up	1652	.37
#Full	up	1733	.39
```

## plot bar chart
```{r, engine='R', eval=F, echo=TRUE}
#module load R/4.1.2 
#R 
library("lattice") 
bar=read.table("overlap_peak.txt", header = TRUE, sep = "\t")
bar$medium<- factor(bar$medium, levels = c("Full", "CS"))
bar$supp<- factor(bar$supp, levels = c("down", "up", "matched"))
bar$group=paste0("ATAC-", bar$medium, "-", bar$supp)
bar$group2=paste0("ATAC-",bar$supp)

pdf('barchart_GATA3_ATAC_overlap_ChIP.pdf', width=10,height=6)
print(barchart(fraction ~ group2 | medium,         
         data = bar,
         groups = supp,
         #stack = TRUE,
         #auto.key=list(space="right"),
         scales = list(x = list(rot = 45)),
         ylim=c(0,1),
         col=c("lightblue", "pink", "white"),
         ylab = "Fraction",
         xlab = "Category",
         main="fraction of the up/down/matched ATAC-seq peaks that overlap GATA3 ChIP peaks"
         #par.settings = my.settings)
))
dev.off()


library("lattice") 
bar=read.table("overlap_peak.txt", header = TRUE, sep = "\t")
bar$medium<- factor(bar$medium, levels = c("Full", "CS"))
bar$supp<- factor(bar$supp, levels = c("down", "matched_Down", "up", "matched_Up"))
bar$group=paste0("ATAC-", bar$medium, "-", bar$supp)
bar$group2=paste0("ATAC-",bar$supp)

pdf('barchart_GATA3_ATAC_overlap_ChIP.pdf', width=10,height=6)
print(barchart(fraction ~ group2 | medium,         
         data = bar,
         groups = supp,
         #stack = TRUE,
         #auto.key=list(space="right"),
         scales = list(x = list(rot = 45)),
         ylim=c(0,1),
         col=c("lightblue", "white", "pink", "white"),
         ylab = "Fraction",
         xlab = "Category",
         main="fraction of the up/down/matched ATAC-seq peaks that overlap GATA3 ChIP peaks"
         #par.settings = my.settings)
))
dev.off()


# Plot the Full data set
library("lattice") 
bar=read.table("overlap_peak.txt", header = TRUE, sep = "\t")[5:8,]
bar$fraction2=1-bar$fraction
overlap_df=bar[, c(1,2,4)]
overlap_df$supp2="overlap"
nonoverlap_df=bar[, c(1,2,5)]
nonoverlap_df$supp2="non-overlap"
colnames(nonoverlap_df)=c("medium", "supp", "fraction", "supp2" )

df=rbind(overlap_df, nonoverlap_df)
df$supp=factor(df$supp, levels=c("down", "matched_Down", "up", "matched_Up"))
df$supp2=factor(df$supp2, levels=c("overlap", "non-overlap"))
df$supp3=rep(c("Decreased ATAC", "Matched with Decreased", "Matched with Increased", "Increased ATAC"), 2)
df$supp3=factor(df$supp3, levels=c("Decreased ATAC", "Matched with Decreased", "Increased ATAC", "Matched with Increased"))

pdf("stack_barchart_GATA3_Full_ATAC_overlap_ChIP.pdf", width=10, height=6)
print(barchart(fraction ~ supp3, 
               data = df, 
               groups = supp2, 
               scales = list(x = list(rot = 45, col=c("#2290cf", "grey20", "#ce228e", "grey20"), font=2)),
               between=list(y=1.0, x = 1.0),
               auto.key = list(points = FALSE, rectangles = TRUE, space = "right"),
               aspect = 1,
               stack = TRUE, 
               col = c("black", "white"),
               ylab = "Fraction of ATAC-seq overlapping ChIP", 
               xlab = "",
               main="Cell Cultured with Full Medium",
               par.settings = list(superpose.polygon = list(col=c("black", "white"), pch=20, lwd=2), strip.background=list(col="grey85"))
               )
)
dev.off()

# plot the ChIP data set 
library("lattice") 
bar=read.table("overlap_peak.txt", header = TRUE, sep = "\t")[1:4,]
bar$fraction2=1-bar$fraction
overlap_df=bar[, c(1,2,4)]
overlap_df$supp2="overlap"
nonoverlap_df=bar[, c(1,2,5)]
nonoverlap_df$supp2="nonoverlap"
colnames(nonoverlap_df)=c("medium", "supp", "fraction", "supp2" )


df=rbind(overlap_df, nonoverlap_df)
df$supp=factor(df$supp, levels=c("down", "matched_Down", "up", "matched_Up"))
df$supp2=factor(df$supp2, levels=c("overlap", "nonoverlap"))
df$supp3=rep(c("Decreased ATAC", "Matched with Decreased", "Matched with Increased", "Increased ATAC"), 2)
df$supp3=factor(df$supp3, levels=c("Decreased ATAC", "Matched with Decreased", "Increased ATAC", "Matched with Increased"))


pdf("stack_barchart_GATA3_ATAC_CS_overlap_ChIP.pdf", width=10, height=6)
print(barchart(fraction ~ supp3, 
               data = df, 
               groups = supp2, 
               scales = list(x = list(rot = 45, col=c("#2290cf", "grey20", "#ce228e", "grey20"), font=2)),
               between=list(y=1.0, x = 1.0),
               auto.key = list(points = FALSE, rectangles = TRUE, space = "right"),
               aspect = 1,
               stack = TRUE, 
               col = c("black", "white"),
               ylab = "Fraction of ATAC-seq overlapping ChIP", 
               xlab = "",
               main="Cell Cultured with Charcoal-stripped Medium",
               par.settings = list(superpose.polygon = list(col=c("black", "white"), pch=20, lwd=2), strip.background=list(col="grey85"))
)
)
dev.off()

```


# youden Index (Y.I)

The goal of this analysis is answer a question: Which GATA3 ChIP-seq peaks functionally regulate chromatin accessibility? \
We can empirically determining a threshold that we can use that as the lower bound of a potentially functional peak. \

Youden Index helps to find the optimal cut-off that maximizes the true positive rate (TPR) while minimizing the false positive rate (FPR), ultimately maximizing the Youden Index (Youden Index = TPR − FPR). \

We can calculate the Youden Index by using the Receiver Operating Characteristic (ROC) curve: \

1. For each possible cut-off point, calculate the true positive rate (TPR) and false positive rate (FPR). \
2. The TPR is the proportion of Group 2 (positive class: intensity of ChIP-seq peaks overlapping with ATAC-seq decreased peaks) correctly identified as positive. \
3. The FPR is the proportion of Group 1 (negative class: intensity of ChIP-seq peaks overlapping with "matched-ATAC-seq decreased" – unchanged ATAC-seq peaks) incorrectly identified as positive. \

Receiver-operating characteristic curve (ROC): \

1. The x-axis represents the False Positive Rate (FPR). \
2. The y-axis represents the True Positive Rate (TPR). \
3. A perfect classifier would have a curve that reaches the top-left corner of the plot, meaning a high TPR with a low FPR. \

4. The diagonal line from the bottom-left to the top-right represents a random classifier (no predictive power). \
The area under the ROC curve (AUC) is often used as a metric to summarize the model’s performance. An AUC of 1 represents a perfect model, while an AUC of 0.5 indicates a model with no discriminatory power (random guessing). \

7023 CS_GATA3_degron_ATAC_down_200bp_sorted.bed
7023 CS_GATA3_degron_ATAC_matched_down_200bp_sorted.bed
  
14057 Full_GATA3_degron_ATAC_down_200bp_sorted.bed
14057 Full_GATA3_degron_ATAC_matched_down_200bp_sorted.bed

## get ATAC-seq peaks (in different categories: down/matched-down) that overlap with GATA3 ChIP-seq peaks (with the ChIP-seq peak intensity). 

We use the `intersectBed` tool to get overlapped peaks between 200bp window ATAC-seq regions and 200bp window ChIP-seq regions: \

```{bash}
module load bedtools
ChIP_file=/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/peak_call/GATA_ChIP_summit_200window.bed #96868

for i in /home/FCAM/ssun/GATA3_ATAC_July2024/ATAC_analysis/overlap3_atac_overlap_with_chip/Full*down*_200bp_sorted.bed # down ATAC peaks and matched-down ATAC peaks, both 14057
do
  nm=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_200bp_sorted.bed" '{print $1}')
  echo $nm
  intersectBed -v -wa -a $ChIP_file -b $i | intersectBed -v -wa -a $ChIP_file -b - -f 1 > ${nm}_overlap_with_ChIP.bed
done

for i in /home/FCAM/ssun/GATA3_ATAC_July2024/ATAC_analysis/overlap3_atac_overlap_with_chip/CS*down*_200bp_sorted.bed
do
  nm=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_200bp_sorted.bed" '{print $1}')
  echo $nm
  intersectBed -v -wa -a $ChIP_file -b $i | intersectBed -v -wa -a $ChIP_file -b - -f 1 > ${nm}_overlap_with_ChIP.bed
done

# These are ChIP-seq peaks!!!!!!!!!!!!!!!
#6598 CS_GATA3_degron_ATAC_down_overlap_with_ChIP.bed
#2114 CS_GATA3_degron_ATAC_matched_down_overlap_with_ChIP.bed
#12751 Full_GATA3_degron_ATAC_down_overlap_with_ChIP.bed
#3287 Full_GATA3_degron_ATAC_matched_down_overlap_with_ChIP.bed


# below code similar to above except that without the '-f 1 ' flag -- this flag sets the minimum fraction of the A feature that must be overlapped by B
for i in /home/FCAM/ssun/GATA3_ATAC_July2024/ATAC_analysis/overlap3_atac_overlap_with_chip/Full*down*_200bp_sorted.bed
do
  nm=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_200bp_sorted.bed" '{print $1}')
  echo $nm
  intersectBed -v -wa -a $ChIP_file -b $i | intersectBed -v -wa -a $ChIP_file -b - > ${nm}_overlap_with_ChIP.bed
done

for i in /home/FCAM/ssun/GATA3_ATAC_July2024/ATAC_analysis/overlap3_atac_overlap_with_chip/CS*down*_200bp_sorted.bed
do
  nm=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_200bp_sorted.bed" '{print $1}')
  echo $nm
  intersectBed -v -wa -a $ChIP_file -b $i | intersectBed -v -wa -a $ChIP_file -b - > ${nm}_overlap_with_ChIP.bed
done

# below are coherence check
ChIP_file=/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/peak_call/GATA_ChIP_summit_200window.bed
i=Full_GATA3_degron_ATAC_down_200bp_sorted.bed
intersectBed -v -wa -a $ChIP_file -b $i | wc -l #84117 --- ChIP that do not overlap with ATAC
intersectBed -v -wa -a $i -b $ChIP_file | wc -l #1032 --- ATAC that do not overlap with ChIP

intersectBed -v -wa -a $ChIP_file -b $i | intersectBed -v -wa -a $ChIP_file -b - -f 1 | wc -l #12751 (96868-84117)

intersectBed -v -wa -wb -a $i -b $ChIP_file | intersectBed -v -wa -wb -a $i -b - -f 1 | wc -l #13025 # ATAC-seq

intersectBed -v -a $i -b $ChIP_file | intersectBed -v -a $i -b - -f 1 | intersectBed -wa -a $ChIP_file -b - | wc -l #13118 # ChIP-seq
intersectBed -v -a $i -b $ChIP_file | intersectBed -v -a $i -b - -f 1 | intersectBed -wa -wb -a $ChIP_file -b - | uniq | wc -l #13118 there are more than 1 atac seq peaks overlap with the same chip seq peaks
intersectBed -v -a $i -b $ChIP_file | intersectBed -v -a $i -b - -f 1 | intersectBed -wa -a $ChIP_file -b - | uniq | wc -l #12751
intersectBed -v -a $i -b $ChIP_file | intersectBed -v -a $i -b - -f 1 | intersectBed -wa -wb -a - -b $ChIP_file | uniq


# coherence check
intersectBed -v -a $i -b $ChIP_file | intersectBed -v -a $i -b - -f 1 | intersectBed -wa -a $ChIP_file -b - | sort -k1,1 -k2,2n | uniq --count --repeated | wc -l # 367
#13118-12751 # ChiP-seq peaks that overlap with duplicated ATAC
#[1] 367


intersectBed -v -a $ChIP_file -b $i  | intersectBed -v -a $ChIP_file -b - -f 1 | intersectBed -wa -a $i -b - | sort -k1,1 -k2,2n | uniq --count --repeated | wc -l # ATAC-seq peaks that overlap with the same ChIP-seq peaks
# 13118-13025
#[1] 93


intersectBed -a $ChIP_file -b $i | sort -k1,1 -k2,2n | uniq --count --repeated | wc -l
#13025-12751  
# 367-93
#[1] 274

```


6598 CS_GATA3_degron_ATAC_down_overlap_with_ChIP.bed
2114 CS_GATA3_degron_ATAC_matched_down_overlap_with_ChIP.bed

12751 Full_GATA3_degron_ATAC_down_overlap_with_ChIP.bed
3287 Full_GATA3_degron_ATAC_matched_down_overlap_with_ChIP.bed

## plot violin plot of the four categories

```{r}
#module load R/4.1.2
#R
path.to.bed="/home/FCAM/ssun/GATA3_ATAC_July2024/ATAC_analysis/overlap3_atac_overlap_with_chip/"
df.overlap.peak = data.frame()     
for (bed in Sys.glob(file.path(paste0(path.to.bed, "*_overlap_with_ChIP.bed")))) {
        factor.name = strsplit(strsplit(bed, "/")[[1]][length(strsplit(bed, "/")[[1]])], '_GATA3_')[[1]][1]
        factor.name1= strsplit((strsplit(strsplit(bed, "/")[[1]][length(strsplit(bed, "/")[[1]])], 'GATA3_degron_')[[1]][2]), "_with_ChIP.bed")[[1]][1]
        print(factor.name)
        print(factor.name1)
        temp = read.table(bed, header=F, comment.char='', col.names=c("V1", "V2", "V3", "V4", "intensity"))
        temp$group=paste0(factor.name, "_", factor.name1)
        df.overlap.peak = rbind(df.overlap.peak, temp)
}

colnames(df.overlap.peak)=c("chr", "start", "end", "V4", "Intensity", "group")

df.overlap.peak$group=factor(df.overlap.peak$group, levels=c( "Full_ATAC_down_overlap", "Full_ATAC_matched_down_overlap", "CS_ATAC_down_overlap", "CS_ATAC_matched_down_overlap"))      

library(lattice)
library(latticeExtra)

pdf('violinplot_ChIP_intensity_overlap_with_ATACdown_or_matcheddown_peaks.pdf',width=10,height=6)
print(bwplot((log(Intensity, base = 10)) ~ group,
             groups = df.overlap.peak$group,
             horizontal = F,
             do.out=FALSE,
             data=df.overlap.peak,
             #ylim=c(-2, 8),
             #groups = motif, 
             #layout=c(1,2),
             #pch="|",
             #col=c("grey", "lightgrey", "lightpink", "lightgrey", "lightpink"),
             panel = function(x, y, ...) {
                                       #panel.stripplot(x, y, jitter.data = TRUE, amount = 0.4, col = adjustcolor("grey", alpha.f = 0.3), pch = 16, cex = 0.3, ...)
                                       panel.violin(x,y, col = "transparent",...)
                                       #panel.abline(h = c(1.98, 1.91), col = "red", lty = 2) # 95 and 82
       },
             #lty = 1,
             
             scales=list(x=list(cex=1,relation = "free", rot = 45), 
                         y =list(cex=1, relation="free")),
             #par.settings = list(superpose.line = list(col=c("grey", "lightgrey", "lightpink", "lightgrey", "lightpink"), lwd=3), strip.background=list(col="grey85")),
             main = "Intensity of ChIP-seq peaks overlapping ATAC-seq down /matched_down peaks",
             ylab = "log10 Intensity"
             )
)
dev.off()
```

## Determine the intensity cut-off for ChIP-seq peaks overlapping with significantly decreased or matched decreased peaks from Full ATAC-seq

```{r}
library(pROC)
library(lattice)

#head(df.overlap.peak) #[1] 24750
full_intensity=rbind(df.overlap.peak[df.overlap.peak$group=="Full_ATAC_down_overlap", ], df.overlap.peak[df.overlap.peak$group=="Full_ATAC_matched_down_overlap", ]) #[1] 16038
full_intensity$group=factor(full_intensity$group, levels=c("Full_ATAC_matched_down_overlap", "Full_ATAC_down_overlap"))
roc_curve <- roc(full_intensity$group, full_intensity$Intensity) #the first level of the factor as the "negative" class (0) and the second level as the "positive" class

#calculate youden index
youden_index <- roc_curve$sensitivities + roc_curve$specificities - 1

# Find the optimal threshold
optimal_idx <- which.max(youden_index)
optimal_threshold <- roc_curve$thresholds[optimal_idx]

cat("Optimal Cut-Off Threshold:", optimal_threshold, "\n")
#Optimal Cut-Off Threshold: 27.5928 
```

```{r}
roc_data <- data.frame(
  FPR = 1 - roc_curve$specificities,
  TPR = roc_curve$sensitivities
)

# Get the optimal FPR and TPR
optimal_FPR <- 1 - roc_curve$specificities[optimal_idx]
optimal_TPR <- roc_curve$sensitivities[optimal_idx]

pdf("Full_ROC_curve.pdf")
print(
  xyplot(TPR ~ FPR, data = roc_data, type = "o", col = "blue",
         xlab = "False Positive Rate (1 - Specificity)",
         ylab = "True Positive Rate (Sensitivity)",
         main = "ROC Curve with Optimal Cut-Off Point (Full)",
         panel = function(x, y, ...) {
           # Plot the ROC curve
           panel.xyplot(x, y, ...)
           
           # Add the optimal point (red dot)
           panel.points(optimal_FPR, optimal_TPR, col = "red", pch = 19, cex = 1.5)
           
           # Add label for the threshold at the optimal point
           panel.text(optimal_FPR, optimal_TPR, labels = round(optimal_threshold, 2), pos = 3, col = "red")
           
           # Add dashed vertical and horizontal lines at the optimal point
           panel.abline(v = optimal_FPR, lty = 2, col = "darkgray")  # Vertical line
           panel.abline(h = optimal_TPR, lty = 2, col = "darkgray")  # Horizontal line
           
           # Print the coordinates (FPR, TPR) next to the lines
           panel.text(optimal_FPR, 0, labels = paste("FPR:", round(optimal_FPR, 2)), pos = 1, col = "darkgray")  # At the bottom of vertical line
           panel.text(1, optimal_TPR, labels = paste("TPR:", round(optimal_TPR, 2)), pos = 2, col = "darkgray")  # At the end of horizontal line
         })
)
dev.off()
```

```{r}
youden_data <- data.frame(
  Threshold = roc_curve$thresholds,
  Youden_Index = youden_index
)
pdf("Full_YoudenIndex_vs_threshold_curve.pdf")
print(
xyplot(Youden_Index ~ Threshold, data = youden_data, type = "o", col = "green",
       xlab = "Threshold",
       ylab = "Youden Index",
       main = "Youden Index vs. Threshold (Full)",
       panel = function(x, y, ...) {
         panel.xyplot(x, y, ...)
         panel.points(optimal_threshold, youden_index[optimal_idx],
                      col = "red", pch = 19, cex = 1.5)
         panel.text(optimal_threshold, youden_index[optimal_idx],
                    labels = round(optimal_threshold, 2),
                    pos = 3, col = "red")
       })
)
dev.off()
```


# Plot the GATA3 binding regions' binding intensity (--filter with Y.I. determined cutoff)

## Full cut-off
Optimal Cut-Off Threshold:  27.5928 ---Full dataset

```{bash}
dir=/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/peak_Intensity/GATA_Deseq2/
awk '$4 > 27.5928' ${dir}GATA_ChIP_summits_final_DESeq2_ranked_intensity.bed > GATA_ChIP_summits_final_DESeq2_ranked_intensity_Fullcutoff.bed #89241 (filterred) of 96868 (raw)

cd /home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/ChIPpeaks_with_ind_motif/compare_binding_intensity_101win_Fullfilter/
awk '$4 < 27.5928' ${dir}GATA_ChIP_summits_final_DESeq2_ranked_intensity.bed > GATA_ChIP_summits_final_DESeq2_ranked_intensity_Fullcutoff_excluded.bed

sizes=/home/FCAM/ssun/Genome/hg38.chrom.sizes
genome=/home/FCAM/ssun/Genome/hg38.fa


# extend the ChIP-peak window size to 101bp
module load bedtools
slopBed -b 50 -i GATA_ChIP_summits_final_DESeq2_ranked_intensity_Fullcutoff.bed -g $sizes  | sort -k1,1 -k2,2n > GATA_ChIP_summits_final_DESeq2_ranked_intensity_Fullcutoff_100window.bed
fastaFromBed -fi $genome -bed GATA_ChIP_summits_final_DESeq2_ranked_intensity_Fullcutoff_100window.bed -fo GATA_ChIP_summits_final_DESeq2_ranked_intensity_Fullcutoff_100window.fasta
```

```{bash}
module load meme/5.4.1
module load bedtools
module load R/4.1.2
for i in individual_meme/*_meme.txt # meme PWMs 
do
    name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_meme.txt" '{print $1}')
    echo $name
    mast -hit_list -best $i GATA_ChIP_summits_final_DESeq2_ranked_intensity_Fullcutoff_100window.fasta > mast_${name}_in_peaks.txt
    mot=`grep MOTIF $i | cut -f 2 `
    echo $mot
    Rscript /home/FCAM/ssun/scripts/updated_parse_mast_to_coordinates.R mast_${name}_in_peaks.txt
    awk -v nm="$name" '{OFS="\t";} {print $0, nm}' mast_${name}_in_peaks.bed | sort -k1,1 -k2,2n > mast_${name}_indiv.bed
done

intersectBed -C -names 10 11 12 1 2 3 4 5 6 7 8 9 -wa -wb -a GATA_ChIP_summits_final_DESeq2_ranked_intensity_Fullcutoff_100window.bed -b *_indiv.bed > x.bed
```


## compare the ChIP-seq peak intensity for regions have different motifs

```{r}
#module load R/4.1.2
#R

library(dplyr)
library(MatchIt)

x = read.table('x.bed', header=FALSE)

result <- x %>%
  group_by(V1, V2, V3) %>% # V2, V3, V4 represent the columns for chromosome, start, and end positions
  summarise(cum_sum = sum(V6)) %>% # V6 represents the column to sum
  filter(cum_sum == 1)  # Filter out groups where the cumulative sum is 1

y <- x %>%
  inner_join(result, by = c("V1" = "V1", "V2" = "V2", "V6" = "cum_sum"))
#nrow(y)
#[1] 30709


colnames(y) = c('chr', 'start', 'end', 'intensity', 'motif', 'num', 'end2')
head(y)
#   chr   start     end intensity motif num    end2
#1 chr1  845746  845847 530.34196    11   1  845847
#2 chr1 1080506 1080607  29.14573     1   1 1080607
#3 chr1 1238187 1238288  43.01128     1   1 1238288
#4 chr1 1504971 1505072  51.35670     6   1 1505072
#5 chr1 1686381 1686482  66.57065     1   1 1686482
#6 chr1 1724297 1724398  34.52345     6   1 1724398
```

```{r}
library(lattice)
library(latticeExtra)
# plot the log2 ChIP-seq intensity

counts <- table(y$motif)
motif_levels <- levels(factor(y$motif))

# plot the ChIP-seq intensity (log scaled)
pdf('bwviolinplot_ChIP_intensity_peaks_101win_with_single_motif_Fullcutoff.pdf', useDingbats = FALSE, width=7.5, height=4.4)

trellis.par.set(box.umbrella = list(lty = 1, col="#93939380", lwd=2),
                box.rectangle = list(col = '#93939380', lwd=1.6),
                plot.symbol = list(col='#93939380', lwd=1.6, pch ='.'))

    print(bwplot(log(y$intensity, base = 2) ~ y$motif,             
             between=list(y=1.0, x = 1.0),
             scales=list(x=list(draw=TRUE),relation="free",rot = 45, alternating=c(1,1,1,1),cex=1,font=1),
             xlab = 'GATA3 motif',
             main = "",
             ylim = c(3,12),
             ylab =expression("log"[2]~"ChIP-seq Intensity"),
             horizontal =FALSE,  
             col= 'black',
             aspect = 0.3,
             par.settings=list(par.xlab.text=list(cex=1.2,font=1),
                               par.ylab.text=list(cex=1.2,font=1),
                               par.main.text=list(cex=1.2, font=1),
                               plot.symbol = list(col='black', lwd=1.6, pch =19, cex = 0.25)),
             strip = function(..., which.panel, bg) {
               #bg.col = c("#ce228e" ,"grey60", "#2290cf","grey90")
               strip.default(..., which.panel = which.panel,
                             bg = rep(bg.col, length = which.panel)[which.panel])
             },
             panel = function(..., box.ratio, col) {
                 panel.abline(h = 0, col = 'grey45', lty = 2)
                 panel.violin(..., col = '#d9c8b1', 
                              varwidth = FALSE, box.ratio = box.ratio, outer = FALSE)
                 #panel.stripplot(..., col='#724ed880', do.out=FALSE, jitter.data=TRUE, amount = 0.2, pch = 16)
                 panel.bwplot(..., pch = '|', do.out = FALSE)
                
                  # Add the "n=xx" text above each violin
               unique_x <- as.numeric(factor(motif_levels))  # Get positions for motif levels
               panel.text(x = unique_x,  # Positions corresponding to the motif groups
                          y = rep(9.5, length(motif_levels)),  # Adjust the y position above the violin
                          labels = paste0("n=", counts), 
                          pos = 3, cex = 0.8, col = 'black')
             }))
dev.off()

# plot the ChIP-seq intensity
pdf('bwviolinplot_ChIP_intensity_peaks_101win_with_single_motif_Fullcutoff2.pdf', useDingbats = FALSE, width=7.5, height=4.4)

trellis.par.set(box.umbrella = list(lty = 1, col="black", lwd=2),
                box.rectangle = list(col = 'black', lwd=1.6),
                plot.symbol = list(col='black', lwd=1.6, pch ='.'))

    print(bwplot(y$intensity ~ y$motif,             
             between=list(y=1.0, x = 1.0),
             scales=list(x=list(draw=TRUE),relation="free",rot = 45, alternating=c(1,1,1,1),cex=1,font=1),
             xlab = 'GATA3 motif',
             main = "",
             #do.out=FALSE,
             ylim = c(0,350),
             ylab ="ChIP-seq Intensity",
             horizontal =FALSE,  
             col= 'black',
             aspect = 0.3,
             par.settings=list(par.xlab.text=list(cex=1.2,font=1),
                               par.ylab.text=list(cex=1.2,font=1),
                               par.main.text=list(cex=1.2, font=1),
                               plot.symbol = list(col='black', lwd=1.6, pch =19, cex = 0.25)),
             strip = function(..., which.panel, bg) {
               #bg.col = c("#ce228e" ,"grey60", "#2290cf","grey90")
               strip.default(..., which.panel = which.panel,
                             bg = rep(bg.col, length = which.panel)[which.panel])
             },
             panel = function(..., box.ratio, col) {
                 panel.abline(h = 0, col = 'black', lty = 2)
                 #panel.violin(..., col = '#d9c8b1', 
                 #             varwidth = FALSE, box.ratio = box.ratio, outer = FALSE)
                 #panel.stripplot(..., col='#724ed880', do.out=FALSE, jitter.data=TRUE, amount = 0.2, pch = 16)
                 panel.bwplot(..., pch = '|', do.out = FALSE)
           
       }))
dev.off()
```



# Plot the GATA3 binding regions' binding intensity (--simple, and use peak stacks in motif stack bar plot)
Instructions: \
1) Take all the peaks from the barchart that have a motif.
2) separate them by their motif category by literally just taking the peaks out of the stacked barplot. Some peaks may have multiple motifs, especially for the early identified motif-containing peaks. It doesn’t matter.
3) take the left over 20% of peaks without motifs.
4) BW plot the ChIP intensities of these categories.

```{r}
cd /home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/ChIPpeaks_with_motif
dir="/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/final_Exhaustive_denovomotif_analysis_240418/"
cp ${dir}*without_motifs_*.bed .
rm ${dir}*without_motifs_*_*.bed
ls
without_motifs_123456789101112.bed  without_motifs_123456.bed
without_motifs_1234567891011.bed    without_motifs_12345.bed
without_motifs_12345678910.bed      without_motifs_1234.bed
without_motifs_123456789.bed        without_motifs_123.bed
without_motifs_12345678.bed         without_motifs_12.bed
without_motifs_1234567.bed          without_motifs_1.bed

head without_motifs_1234567.bed  
chr1	827300	827461	45.324686100056
chr1	924773	924934	30.4468389666785
chr1	966573	966734	26.8701605998383
chr1	999428	999589	21.2245532255625
chr1	1000456	1000617	24.342633538906
chr1	1001811	1001972	25.5287170527213
chr1	1013500	1013661	36.2692161076724
chr1	1020107	1020268	20.6315114686548
chr1	1021062	1021223	32.6995098507957
chr1	1069469	1069630	31.4148299904389


wc -l *.bed
   17319 without_motifs_123456789101112.bed
   17960 without_motifs_1234567891011.bed
   18566 without_motifs_12345678910.bed
   20704 without_motifs_123456789.bed
   22358 without_motifs_12345678.bed
   23994 without_motifs_1234567.bed
   27014 without_motifs_123456.bed
   30822 without_motifs_12345.bed
   35473 without_motifs_1234.bed
   42904 without_motifs_123.bed
   52334 without_motifs_12.bed
   69112 without_motifs_1.bed
  378560 total

head ${dir}GATA_ChIP_summits_final_DESeq2_ranked_intensity_161bpwindow.bed
chr1	827300	827461	45.324686100056
chr1	845716	845877	530.34195637027
chr1	869417	869578	18.626318257943
chr1	916689	916850	36.6999257213148
chr1	917454	917615	124.645017626695
chr1	924773	924934	30.4468389666785
chr1	966573	966734	26.8701605998383
chr1	996069	996230	87.663247085683
chr1	999428	999589	21.2245532255625
chr1	1000456	1000617	24.342633538906
wc -l ${dir}GATA_ChIP_summits_final_DESeq2_ranked_intensity_161bpwindow.bed #96868 

module load bedtools
for i in ${dir}without_motifs_*.bed
do
  nm=$(echo $i | awk -F ${dir}without '{print $2}' | awk -F ".bed" '{print $1}')
  bedtools intersect -a ${dir}GATA_ChIP_summits_final_DESeq2_ranked_intensity_161bpwindow.bed -b without${nm}.bed -v > with${nm}_161win.bed
done

wc -l with_*.bed
   27681 with_motifs_1_161win.bed
   44430 with_motifs_12_161win.bed
   53848 with_motifs_123_161win.bed
   61279 with_motifs_1234_161win.bed
   65933 with_motifs_12345_161win.bed
   69738 with_motifs_123456_161win.bed
   72763 with_motifs_1234567_161win.bed
   74399 with_motifs_12345678_161win.bed
   79448 with_motifs_123456789101112_161win.bed
   78806 with_motifs_1234567891011_161win.bed
   78199 with_motifs_12345678910_161win.bed
   76051 with_motifs_123456789_161win.bed
  782575 total



for i in {2..12}; do
    prev=$((i-1))

    a_file="with_motifs_$(seq -s '' 1 $i)_161win.bed"
    b_file="with_motifs_$(seq -s '' 1 $prev)_161win.bed"
    out_file="with_motifs_${i}_161win.bed"
    bedtools intersect -a "$a_file" -b "$b_file" -v > "$out_file"
done

# coherence check 1
#bedtools intersect -a with_motifs_12_161win.bed -b with_motifs_1_161win.bed -v > with_motifs_2_161win.bed
#bedtools intersect -a with_motifs_123_161win.bed -b with_motifs_12_161win.bed -v > with_motifs_3_161win.bed
#bedtools intersect -a with_motifs_1234_161win.bed -b with_motifs_123_161win.bed -v > with_motifs_4_161win.bed
#bedtools intersect -a with_motifs_12345_161win.bed -b with_motifs_1234_161win.bed -v > with_motifs_5_161win.bed
#bedtools intersect -a with_motifs_123456_161win.bed -b with_motifs_12345_161win.bed -v > with_motifs_6_161win.bed
#bedtools intersect -a with_motifs_1234567_161win.bed -b with_motifs_123456_161win.bed -v > with_motifs_7_161win.bed
#bedtools intersect -a with_motifs_12345678_161win.bed -b with_motifs_1234567_161win.bed -v > with_motifs_8_161win.bed
#bedtools intersect -a with_motifs_123456789_161win.bed -b with_motifs_12345678_161win.bed -v > with_motifs_9_161win.bed
#bedtools intersect -a with_motifs_12345678910_161win.bed -b with_motifs_123456789_161win.bed -v > with_motifs_10_161win.bed
#bedtools intersect -a with_motifs_1234567891011_161win.bed -b with_motifs_12345678910_161win.bed -v > with_motifs_11_161win.bed
#bedtools intersect -a with_motifs_123456789101112_161win.bed -b with_motifs_1234567891011_161win.bed -v > with_motifs_12_161win.bed


# coherence check 2 (17319 peaks without assigned motif)
bedtools intersect -a ${dir}GATA_ChIP_summits_final_DESeq2_ranked_intensity_161bpwindow.bed -b with_motifs_123456789101112_161win.bed -v > without_motifs_test_161win.bed

#compare without_motifs_test_161win.bed vs. without_motifs_123456789101112.bed
intersectBed -v -a without_motifs_test_161win.bed -b without_motifs_123456789101112.bed
intersectBed -v -b without_motifs_test_161win.bed -a without_motifs_123456789101112.bed
intersectBed -a without_motifs_test_161win.bed -b without_motifs_123456789101112.bed | wc -l

cd /home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/ChIPpeaks_with_motif/violin_binding_intensity
wc -l *.bed
   2148 with_motifs_10_161win.bed
    607 with_motifs_11_161win.bed
  27681 with_motifs_1_161win.bed
    642 with_motifs_12_161win.bed
  16749 with_motifs_2_161win.bed
   9417 with_motifs_3_161win.bed
   7431 with_motifs_4_161win.bed
   4654 with_motifs_5_161win.bed
   3805 with_motifs_6_161win.bed
   3025 with_motifs_7_161win.bed
   1636 with_motifs_8_161win.bed
   1652 with_motifs_9_161win.bed
  17419 without_motifs_test_161win.bed
  96866 total

```

```{r}
#module load R/4.1.2
#R

library(dplyr)
library(data.table)
library(stringr)

files <- list.files(pattern = "^with_motifs_([0-9]+)_161win\\.bed$")

dt_list <- lapply(files, function(f) {
  idx <- as.integer(str_match(f, "^with_motifs_([0-9]+)_161win\\.bed$")[,2])
  dt <- fread(f, header = FALSE, sep = "\t",
              col.names = c("chr","start","end","score"))
  dt[, motif := idx]
  setcolorder(dt, c("chr","start","end","score","motif"))  # <- new order
  dt
})

merged_dt <- rbindlist(dt_list)[order(motif, chr, start, end)]

nomotif=read.table('without_motifs_test_161win.bed', header=FALSE)
nomotif$motif="without_motifs"
colnames(nomotif)=c("chr","start","end","score", "motif")

y=rbind(merged_dt, nomotif)
# fwrite(merged_dt, "with_motifs_merged.csv")

head(y)
    chr   start     end     score motif
1: chr1  917454  917615 124.64502     1
2: chr1 1080476 1080637  29.14573     1
3: chr1 1238157 1238318  43.01128     1
4: chr1 1505699 1505860  10.54281     1
5: chr1 1686351 1686512  66.57065     1
6: chr1 1741050 1741211  30.26410     1
> nrow(y)
[1] 96866

```

```{r}
library(lattice)
library(latticeExtra)
# plot the log2 ChIP-seq intensity

counts <- table(y$motif)
motif_levels <- levels(factor(y$motif))

num_levels <- sort(unique(as.integer(as.character(y$motif)[!y$motif %in% "without_motifs"])))
# set the full order with "without_motifs" first
lev <- c("without_motifs", as.character(num_levels))
# re-factor with desired order
y$motif <- factor(as.character(y$motif), levels = lev, ordered = TRUE)

pdf('bwviolinplot_ChIP_intensity_peaks_161win_with_single_motif_250908.pdf', useDingbats = FALSE, width=7.5, height=4.4)

trellis.par.set(box.umbrella = list(lty = 1, col="#93939380", lwd=2),
                box.rectangle = list(col = '#93939380', lwd=1.6),
                plot.symbol = list(col='#93939380', lwd=1.6, pch ='.'))

    print(bwplot(log(y$score, base = 2) ~ y$motif,             
             between=list(y=1.0, x = 1.0),
             scales=list(x=list(draw=TRUE),relation="free",rot = 45, alternating=c(1,1,1,1),cex=1,font=1),
             xlab = 'GATA3 motif',
             main = "",
             ylim = c(3,12),
             ylab =expression("log"[2]~"ChIP-seq Intensity"),
             horizontal =FALSE,  
             col= 'black',
             aspect = 0.3,
             par.settings=list(par.xlab.text=list(cex=1.2,font=1),
                               par.ylab.text=list(cex=1.2,font=1),
                               par.main.text=list(cex=1.2, font=1),
                               plot.symbol = list(col='black', lwd=1.6, pch =19, cex = 0.25)),
             strip = function(..., which.panel, bg) {
               #bg.col = c("#ce228e" ,"grey60", "#2290cf","grey90")
               strip.default(..., which.panel = which.panel,
                             bg = rep(bg.col, length = which.panel)[which.panel])
             },
             panel = function(..., box.ratio, col) {
                 panel.abline(h = 0, col = 'grey45', lty = 2)
                 panel.violin(..., col = '#d9c8b1', 
                              varwidth = FALSE, box.ratio = box.ratio, outer = FALSE)
                 #panel.stripplot(..., col='#724ed880', do.out=FALSE, jitter.data=TRUE, amount = 0.2, pch = 16)
                 panel.bwplot(..., pch = '|', do.out = FALSE)
                
                  # Add the "n=xx" text above each violin
               #unique_x <- as.numeric(factor(motif_levels))  # Get positions for motif levels
               #panel.text(x = unique_x,  # Positions corresponding to the motif groups
                          #y = rep(9.5, length(motif_levels)),  # Adjust the y position above the violin
                          #labels = paste0("n=", counts), 
                          #pos = 3, cex = 0.8, col = 'black')
             }))
dev.off()


```

# Plot the GATA3 binding regions' chromatin accessibility changes

**Plot ATAC-accessibility differences between regions with matched binding intensity and contains different motifs.** \


## find matched (in their chip intensity) ChIP-seq peaks with single motif

### all 12 motifs










